\documentclass[acmsmall, screen]{acmart}
\let\Bbbk\relax

\usepackage{quiver}
\usepackage{mathpartir}
% \usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage{array}

% \theoremstyle{remark} 
% \newtheorem{theorem}{Theorem}
% \newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newenvironment{DIFnomarkup}{}{}
% \pagestyle{empty}
% \settopmatter{printfolios=false}

%\newcommand{\bibliographypath}{../../paper-new/references}
%\newcommand{\defspath}{../../paper-new/defs}
\newcommand{\bibliographypath}{references}
\newcommand{\defspath}{defs}

\newif\ifdraft
\drafttrue
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}


\input{\defspath}


\begin{document}

\title{An Effect-Generic Approach to Proving Operational Graduality}
% \title{Adequacy for Effectful Gradually Typed Languages using Guarded Type Theory}
% \title{Guarded Operational Semantics for Effectful Gradually Typed Languages}
\author{Eric Giovannini}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{ericgio@umich.edu}
\orcid{0009-0003-6871-1714}


\author{Max S. New}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{maxsnew@umich.edu}
\orcid{0000-0001-8141-195X}

\begin{abstract}
    Gradually-typed languages allow for soundly mixing static and
    dynamically typed programming styles and support migration from a
    dynamic to static typing style. A key theorem for such languages
    is graduality, which ensures that the transition from dynamic to
    static typing preserves semantics up to introduction of stricter
    checking. Establishing this property is a metatheoretic challenge,
    and even the correct formulation of this property must be given
    anew for languages with different observable effects.

    In this work we establish a generic graduality theorem for a
    gradually typed lambda calculus parameterized by an algebraic
    theory describing the effects available in the program. Each
    instantiation of this theory provides a concrete operational and
    denotational semantics that generalizes . We then provide a
    adequacy theorem for this denotational semantics and generic
    graduality theorem. The foundations of the framework are built in
    guarded type theory, extending prior work on denotational
    graduality. We show how the generic framework can be instantiated
    to produce many concrete examples of effects: a generic
    trace-based graduality theorem for interaction trees, as well as
    monad transformer theorems that allow us to combine standard
    semantics of state and exception-like effects in a compositional
    manner to produce new instantiations.

    % In this work, we define an operational semantics for a gradually-typed cast
    % calculus parameterized by a collection of algebriac effects. Our main result
    % is a graduality theorem for the operational semantics. To prove this result, we...

    % In this work, we build on the techniques of Giovannini et al, using
    % guarded type theory to prove graduality of an operational semantics for a
    % gradually-typed cast calculus with algebraic effects

    % In this work, we use guarded type theory to prove graduality of an
    % operational semantics for a gradually-typed cast calculus with algebraic
    % effects. We first define a guarded denotational semantics following the
    % techinques of Giovannini et al. Our denotational semantics satisfies a
    % guarded version of graduality; we then connect this result back to the
    % operational semantics via an adequacy theorem. Lastly, we consider special
    % cases in which our result globalizes to the setting of ordinary set theory,
    % obtaining for closed terms of base type a coinductive semantics that
    % satisfies graduality.
\end{abstract}

\maketitle

\newif\ifdraft
\drafttrue
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\eric}[1]{\ifdraft{\color{orange}[{\bf Eric}: #1]}\fi}

% Question: should we present algebraic effects as having a request and response
% type, or should we fold the request type into the type of operations?

\section{Introduction}

\emph{Gradually typed languages} \cite{siek-taha06, tobin-hochstadt06} allow for
safe interoperability between static and dynamic typing disciplines in the same
codebase, and support the smooth, \emph{gradual} migration from dynamic
typing to static typing via the addition of explicit typing information.
without needing to rewrite the entire project in a completely different language.
%
This allows gradually typed programs to support mixed states where
only part of the code has received explicit static typing information,
and developers can gradually attain the benefits of static type
information.
%
One of the fundamental theorems for gradually typed languages
formalizing this idea is \emph{graduality}, also known as the
\emph{dynamic gradual guarantee}, originally defined by Siek,
Vitousek, Cimini, and Boyland \cite{siek_et_al:LIPIcs:2015:5031,
  new-ahmed2018}.
%
Informally, graduality says that migrating code from dynamic to static
typing should only allow for the introduction of static or dynamic
type errors from the stricter checking, and not otherwise change the
observable behavior of the program.

More formally, consider a simple gradually typed language where
programs can diverge or raise a dynamic type error but have no other
visible effects. Then we define a syntactic precision ordering on
terms $M \sqsubseteq N$ that says that $M$ is syntactically identical
to $N$ other than containing more statically precise type
information. Then the graduality theorem says that if $M, N$ are
closed programs with a simple output type such as integers, then
either $M$ and $N$ have the same behavior (both diverge, both
terminate with the same result, both error) or $M$ raises a dynamic
type error.

Despite being a conceptually simple idea, graduality has been a challenge to
prove. The main approaches in prior work are (1) tedious operational simulation
results \cite{siek_et_al:LIPIcs:2015:5031, siek-chen2021} (2) fitting the
language into a particular framework such as Abstracting Gradual Typing or the
Gradualizer \cite{garcia-clark-tanter2016, cimini-siek2016} or (3) operational
logical-relations based methods \cite{new-ahmed2018, new-jamner-ahmed19,
new-licata-ahmed2019} or (4) denotational methods \cite{new-licata18,gtt-sgdt}.
Among these, the advantage of the denotational approach is a decoupling of the
mathematical arguments to particular syntactic details, so that the lemmas that
are core to the graduality proof can be reused modularly across different
developments. A downside of the denotational approach in prior work has been
that it only proves a graduality statement about a denotational semantics, and
so does not prove the same theorem exactly as prior graduality results, which
are defined over operational semantics.

One of the main challenges in developing a graduality proof for a new
language is that even the statement of the theorem itself is sensitive
to the details of what effects are present. For instance if we were to
extend the language above to have (1) system calls to print to stdout
(2) exceptions distinct from the dynamic type error and (3) some
global mutable variables, then we would need to accommodate these in
our theorem. With work the experienced semanticist can come up with a
theorem such as: for any initial state of the global variables, $M$
and $N$, when executed with those variables, make a sequence of
matching system calls up until the point that either $M$ terminates
with a dynamic type error (at which point $N$ can continue in any
manner), they both terminate with the same exceptions, or they both
return the same value with matching outputs in the final mutable
state. This then has to be adjusted for the effects present in
whatever language is being modeled, and the proofs need to be redone
with the appropriate notion of reduction for that system and
observable behavior. The result is a tedious recapitulation of prior
arguments adapted to new effects, but where fundamentally very little
of the argument for graduality truly has changed.

In this work, we develop a framework for proving graduality that
abstracts the concrete details of what effects are present so that the
above systems, and many other combinations can all be developed as
particular instantiations. The core idea of the work is to abstract
over different concrete effects using \emph{algebraic theories}, an
idea commonly known as \emph{algebraic effects} \cite{PLOTKIN2001332, plotkinpower2003}. We
define a gradual lambda calculus whose effects are given by an
algebraic theory that is a parameter to the system. By instantiating
the system with a particular theory and implementation of the free
algebras of the theory, our framework produces (1) an operational
semantics (2) a denotational semantics (3) an adequacy theorem
relating operational and denotational semantics (4) a graduality proof
for the denotational semantics, which is then transported via adequacy
to the operational semantics. These proofs require only that the given
algebraic effect can be lifted to relations, the concrete details of
the use of those relations are then used modularly in the graduality
and adequacy proofs.

The foundation for our work is \emph{guarded type theory} \cite{birkedal-mogelberg-schwinghammer-stovring2011}, which allows
us to define the semantics, adequacy and graduality proofs in a
concise way. Guarded type theory extends dependent type theory with a
later modality and axiomatizes that all guarded recursive definitions
have a fixed point. This provides a concise alternative to the
explicit use of step-indexing\cite{}. Guarded type theory can be
interpreted in the topos of trees, and so any results proven can be
interpreted as ordinary step-indexed mathematical results. It requires
work, however to bridge the gap from the guarded proof of graduality
to the standard results, a process we call \emph{globalization}. To
make our final results as standard as possible, we work through
concrete descriptions of the resulting graduality
results. Specifically, we provide two kinds of constructions. First, we
give a construction for uninterpreted effects like operating system
calls as a kind of trace semantics over coinductively defined
interaction trees \cite{trace,itrees}. Second, we give a
monad-transformer-like construction for state and exception-like
effects. With the interaction trees as a base case, the monad
transformers can be iterated to construct new models on demand.

%% The tradeoff is that results in
%% guarded type theory are constructions 

%% Establishing graduality for a given cast calculus has traditionally been a
%% tedious endeavor. One common approach involves defining a step-indexed logical
%% relation over an operational semantics \cite{new-ahmed2018,
%% new-licata-ahmed2019, new-jamner-ahmed19, new-giovannini-licata-2022}; these
%% methods are inherently dependent on the particular calculus being modeled and
%% offer little opportunity for reuse of mathematical abstractions across different
%% developments. Another method, developed by New and Licata \cite{new-licata18},
%% constructs a denotational semantics using \emph{classical} domain theory. While
%% compositional, this approach suffers from a lack of expressiveness: classical
%% domain-theoretic techniques appear incapable of modeling viciously
%% self-recursive structures such as higher-order store
%% \cite{Birkedal-Stovring-Thamsborg-2009}, which are common in real-world
%% gradually-typed languages.
%% %
%% Thus, it remained an open question whether one could give a semantics for
%% gradual typing that combined the expressiveness of the step-indexed approach
%% with the compositional nature of denotational semantics.

% Graduality has remained a difficult property to establish, with existing
% methods being largely tied to syntactic or operational specifics of the
% language being modeled and offering little opportunity for reuse of common
% mathematical abstractions.

% Recently, Giovannini et al. \cite{gtt-sgdt} combined the expressiveness of the
% step-indexed approach with the compositional nature of denotational semantics
% by...

%% Recently, Giovannini et al. \cite{gtt-sgdt} answered this question affirmatively
%% by constructing a denotational semantics for a simple gradually-typed
%% programming language using the tools of guarded type theory
%% \cite{birkedal-mogelberg-schwinghammer-stovring2011}, and proved that their
%% model validates the graduality property. Their model combines the expressive
%% power of step-indexing techniques with the compositional nature of denotational
%% semantics.
%% %
%% However, the semantics developed by Giovannini et al. has the limitation that it
%% is a single model for a quite bare-bones gradually-typed language. Real-world
%% languages can include effects, polymorphism, dependent types, etc.
%% %
%% Furthermore, it is unclear how to relate their model to a more typical
%% operational semantics for a cast calculus. It is desirable to establish an
%% adequacy result relating their denotational semantics to an operational model.
%% More specifically, writing $\sem{M}$ for the denotation of a term $M$, adequacy
%% states that an equality of denotations $\sem{M} = \sem{N}$ implies that $M$ and
%% $N$ behave the same operationally. This allows one to use facts proved in the
%% denotational semantics to reason about the operational behavior of terms.
%% %
%% More specifically, an adequacy result provides a means of translating the
%% graduality theorem --- a statement about the \emph{denotations} of terms ---
%% into a statement about the \emph{operational behavior} of terms.

% In this work, we seek to address these limitations by
% In this work, we seek to address the limitations of the prior model by 

%% In this work, we take a step towards modeling more realistic gradually-typed
%% languages using guarded type theory: we extend the denotational semantics of
%% Giovannini et al. to account for arbitrary algebraic effects, and we establish
%% an adequacy theorem relating the resulting denotational model to an effectful
%% operational semantics.

% The level of generality needed to model arbitrary algebraic effects means that
% the semantics we produce is correspondingly generic. This is true both
% denotationally and operationally. On the denotational side, we generalize the
% semantics of Giovannini et al. to allow for arbitrary algebraic effects. The
% (guarded) graduality theorem we obtain is essentially a statement relating term
% precision to the behavior of the interaction trees that the terms denote.

% TODO: update
% Likewise, we construct a generic effectful operational semantics for the cast
% calculus in guarded type theory.

% While this approach is applicable to any
% algebraic theory representing the programming language's effects, what is
% missing is a way to relate the small-step operational semantics it produces to a
% more traditional effectful small-step semantics. For example, a small-step
% semantics for a language with global state would normally involve a transition
% relation between pairs of a term and state.

% TODO: update
% The missing step in both the operational and denotational semantics is supplied
% by a suitable notion of algebra that interprets the guarded interaction trees as
% a concrete element of an ``observation'' type.

\subsection{Overview of Approach}\label{sec:overview-of-approach}

The main contribution of this paper is a parameterized effectful
gradually-typed cast calculus with operational semantics satisfying graduality.

The framework is parameterized by an \emph{algebraic theory}, which
specifies what effectful operations are available (e.g., reading from a
state, writing to a state, printing, raising an error) and what
equations are expected to hold (e.g., that the state does not change
between consecutive reads). More details on algebraic theories are
given in Section \ref{sec:modeling-effects}. The parameterized cast
calculus extends a standard gradual cast calculus with terms
corresponding to each of the effect operations present in the provided
algebraic theory $\calA$, giving the programmer access to these
effects.

Every algebraic theory $\calA$ induces a monad, called the
\emph{free-model monad} $T_\calA$ for which $T_\calA(X)$ is the free
algebra of the theory generated by $X$. For particular theories like
state and exceptions, these monads are given by the familiar state and
exception monads \cite{state-exn-thy-monad}. We define a small-step
operational semantics and this induces a multi-step execution
semantics:
%
\[ \run : (\term{A} \to T_{\calAsb}\val{A}), \]
%
where $\calAsb$ is the input theory extended with operations
corresponding to runtime errors and explicit operational reduction steps, marked as
either ``bureaucratic'' reductions or non-well-founded steps. The result is
a coinductive monad $T_{\calAsb}$ that combines the possibility of
diverging with an infinite sequence of non-bureaucratic steps with the
effect operations of the input theory. For particular theories we can
extract a concrete description of this monad $T_{\calAsb}$ and confirm
that this is the usual sort of multi-step execution semantics that we
would define by hand.

The graduality property says that if a term $M$ is syntactically more
precise than a term $N$, then the behavior of $M$ when run is a
refinement of the behavior of $N$. More formally, graduality says that
for any closed terms $M, N$ of type $\nat$, if $M \ltdyn N$, then
$\run{M} \lesssim \run{N}$. Here, $\lesssim$ is an error ordering on
monadic values capturing when two computations behave the same up-to
the left side terminating with an error. To perform the full
graduality proof, we require also an extension of the free algebra
monad for the theory from sets to act on the error ordering and
bismilarity relations we use in the proof.

The final proof of graduality results from the composition of several constructions.
% We leverage the existing denotational model of gradual typing of Giovannini et
% al. \cite{gtt-sgdt}.
%
%\begin{enumerate}
% First, we develop a guarded denotational model of gradual typing with
% effects, taking as inspiration the existing model of Giovannini et al.
First, we develop a guarded denotational model of gradual typing with effects
based on the existing model of Giovannini et al. The result is a \textbf{guarded
denotational graduality} theorem:
%
\begin{theorem}[Guarded Denotational Graduality]
  If $\cdot \vdash M \ltdyn N: \nat$, then $\sem{M} \ltbisim \sem{N}$.
\end{theorem}
%
Here, $x \ltbisim y$ means that computation $x$ is the same as $y$ up to $x$
being an error, and $x$ and $y$ may take a different number of steps.

Next, we relate the operational semantics to the denotational semantics
via an \textbf{adequacy theorem}:
%
\begin{theorem}[Adequacy]
  Let $\cdot \vdash M : \nat$. Then we have $\run{M} \equiv_b \sem{M}$.
\end{theorem}
%
Here, $x \equiv_b y$ relates elements belonging to two different
monads, where the monad used in $\run{M}$ supports the
``bureaucratic'' stepping operation while the monad used in $\sem{M}$
only contains the ``true'' steps. This is to account for the fact that
bureaucratic steps correspond to operational rules such as beta
reductions that are equalities in the denotational semantics.
% The reason for this discrepancy will be explained in the section on
% operational semantics, but for now it is sufficient to note that
The relation $x \equiv_b y$ means that $x$ and $y$ are the same up-to
a \emph{finite} number of bureaucratic steps in $x$. The adequacy
theorem then states that the denotational semantics is in lock-step
with the operational semantics on true steps, but the operational
semantics may contain finite sequences of bureaucratic steps. As a
corollary, this proves that any sequence of bureaucratic steps must
eventually terminate. In order for this strict lock-step equivalence
to hold, we needed to modify the denotational semantics slightly from prior work.
% running a term $M$ does not result in an infinite sequence of bureaucratic
% steps.\max{I think we can just say something specific here like ``bureaucratic
% steps are things like beta reductions that are invisible in the denotational
% semantics''}
With this adjustment, the adequacy theorem is proven as usual by a logical relations
argument, given in Section \ref{sec:relating-the-semantics}.

Finally, we want to simplify the combination of adequacy and
denotational graduality into the simple statement about operational
semantics that we would write by hand. To do this we require one final
step, which is to map our results in guarded type theory to an
ordinary statement about the coinductive semantics.  This
translation process is called \emph{globalization} \cite{TODO}. The
motivation is that the statements of the results up to this point
involve features of guarded type theory, e.g., the later modality. The
technique of globalization produces a semantics that does not rely on
any features of guarded type theory, but instead makes sense in
standard type or set theoretic foundations.
%
This step requires some further effect theory-specific technical
conditions in order to produce the final operational graduality
theorem.


We package the input data for our construction above into a ``graduality
schema'' consisting of an algebraic theory, a lifting of the free algebras to
act on relations, and the final global ordering relations. In order to show the
applicability of our framework, we give two methods for constructing graduality
schemata.

First, for a \emph{free} theory, meaning a theory with operations but
no equations, the semantics can be understood as an
\emph{interaction tree}, and the graduality theorem can be formalized
as a concrete relation on finite execution traces of the trees,
generalizing prior work that used an ordering on the delay monad.
%
Secondly, many familiar effect theories can be constructed from
\emph{theory transformers} and their monads from corresponding
\emph{monad transformers}, which are typically much simpler than a
trace semantics. We provide several examples implementing graduality
schemata for common monad transformers. Then starting from an
interaction tree base case, we can build up a complex effect theory
with a concrete operational graduality result.

The organization of the remainder of the paper is as follows:
\begin{enumerate}
    \item In Section \ref{sec:modeling-effects}, we provide background
      on guarded type theory and algebraic theories that will be used
      in the remainder of the paper.
    \item In Section \ref{sec:syntax}, we specify the syntax for the cast
    calculus and definitions of type and term precision used to define the
    graduality property.
    \item In Section \ref{sec:operational-semantics}, we define the operational
    semantics for the cast calculus defined in the previous section.
    \item In Section \ref{sec:denotational-semantics}, we show how to
      construct a denotational semantics that is ``gradual by
      construction'' for the effectful cast calculus by adapting the
      construction of Giovannini et al. \cite{gtt-sgdt} to account for
      arbitrary algebraic effects.
    \item In Section \ref{sec:relating-the-semantics}, we prove an adequacy
    theorem relating the operational and denotational models in the expected
    manner.
    \item In Section \ref{sec:globalization}, we discuss how to globalize the
    results of the previous two sections.
    \item Lastly, in Section \ref{sec:discussion}, we summarize our results,
    compare to prior work, and conclude with directions for future work.
\end{enumerate}

\section{Background: Guarded Type Theory and Algebraic Effects}

\subsection{Metatheoretic Preliminaries}

We begin with a brief discussion of the ambient guarded type theory in which we
work. \emph{Guarded type theory} refers to a class of type theories with
primitives that allow one to solve non-well-founded recursive equations and
reason about such solutions. It can be viewed a \emph{synthetic} approach to
step-indexed techniques \cite{birkedal-mogelberg-schwinghammer-stovring2011};
the benefit of this approach is that we avoid the tedious natural number
manipulations present in explicit step-indexing.

There are many variants of guarded type theory, but they each allow a form of
recursion to be encoded at the level of types through a modality $\later \colon
\type \to \type$ (pronounced ``later''). The use of a modality to express
guarded recursion originates in the work of Nakano \cite{Nakano2000}. More
specifically, a typical guarded type theory has at least the following features:

\begin{itemize}
  \item Given a type $A$, the type $\later A$ represents an element of type $A$
  that is available one ``abstract time step'' later. 

  \item There is an operator $\nxt : A \to\, \later A$ that ``delays'' an element
  available now to make it available later.

  \item There is a principle of \emph{guarded recursion}, implemented by a
  guarded fixpoint operator of the form
  %
  \[ \fix : \forall A, (\later A \to A) \to A. \] 
  %
  That is, to construct a term of type $A$, it suffices to assume that we have
  access to such a term ``later'' and use that to help us build a term ``now''.
  This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$.

  \item There is a method for solving guarded-recursive type equations $T \cong
  F(\later T)$. In guarded type theory with a universe of types $\type$, we can
  solve such equations by applying the guarded fixpoint operator with $A =
  \type$ \cite{birkedal-møgelberg-2013}.
  %
  In particular, $\fix$ applies when type $T$ is instantiated to a proposition
  $P : \texttt{Prop}$; in that case, it corresponds to $\lob$-induction.
  
\end{itemize}
 
We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.

% TODO later is an applicative functor, but not a monad

% Clocked Cubical Type Theory
\subsubsection{Ticked Cubical Type Theory}

Following Giovannini et al. \cite{gtt-sgdt}, in this paper we will work in a
specific form of guarded type theory with access to additional constructs called
\emph{ticks} and \emph{clocks}.

% TODO review and consider trimming
The specific type theory in which we work is called Ticked Cubical Type Theory
\cite{mogelberg-veltri2019}, an extension of Cubical Type Theory
\cite{CohenCoquandHuberMortberg2017} that has an additional sort called
\emph{ticks}. Ticks were originally introduced by Bahr et al.
\cite{bahr-grathwohl-bugge-mogelberg2017}. A tick $t : \tick$ intuitively acts
as evidence that one unit of time has passed. In Ticked Cubical Type Theory, the
type $\later A$ is the type of dependent functions from ticks to $A$. The type
$A$ is allowed to depend on $t$, in which case we write $\later_t A$ to
emphasize the dependence.
%
For the sake of brevity, we will also write tick application as $M_t$. The
constructions we carry out in this paper will take place in a guarded type
theory with ticks.

\section{Modeling Effects with Algebraic Theories}\label{sec:modeling-effects}

In this section, we discuss our approach to modeling algebraic effects in the
context of gradual typing. These concepts will be employed in both our
denotational and operational semantics.

\subsection{Guarded Algebraic Theories}\label{sec:guarded-algebraic-theories}

We introduce the notion of a \emph{guarded algebraic theory}, a suitable
generalization of the notion of algebraic theory where we allow for the
algebraic operations to have their arguments guarded by the later modality. We
begin with the notion of a guarded algebraic signature.
% TODO cite paper on algebraic effects

A \textbf{guarded algebraic signature} $\Sigma$ consists of a collection of
operation symbols $\Op$. Each operation symbol $\oper : \Op$ has an
\textbf{ordinary arity} given by a type $\tau_\oper$ and a \textbf{guarded
arity} given by a type $\gamma_\oper$. We write $\oper : (\tau, \gamma)$ to mean
$\oper$ has ordinary arity $\tau$ and guarded arity $\gamma$.
% Each operation \oper : (\tau, \gamma) has ordinary arity \tau and guarded arity \gamma.
%
% We do not require the arity types to be finite.
If the arity types are finite, i.e., $\tau_\oper = [n]$ for some $n$ with $[n]$
denoting the finite type with $n$ elements, and likewise for $\gamma_\oper$,
then we say that the signature is \textbf{finitary}.

\begin{remark}
  In defining algebraic effects, one typically defines an algebraic signature as
  a collection of operations that each have a \emph{request} type $\sigma_\oper$
  and response type $\tau_\oper$. The request type of operation $op$ is the type
  of the value that the program sends to the environment. Similarly, the
  \emph{response type} of the operation is the type of the value that the
  environment sends back to the program once the operation is complete. We can
  encode this definition into ours by having a separate algebraic operation
  $\oper_x$ for each $x : \sigma$.
\end{remark} 

Given a guarded signature $\Sigma$ as described above, we define the
\textbf{polynomial endofunctor} corresponding to $\Sigma$: 
%
\[ \Sigma(X) = \Sigma_{\oper: \Op}, (\tau_\oper \to X) \times (\gamma_\oper \to \later X). \]
%

A \textbf{$\Sigma$-algebra} is simply an algebra of the endofunctor $\Sigma$
defined above, i.e, a type $B$ equipped with a function $\Sigma(B) \to B$. More
concretely, a $\Sigma$-algebra consists of a set $B$ equipped with
interpretations for each of the operations in $\Sigma$. That is, for each $\oper
: (\tau, \gamma)$, there is a function
%
\[ \sem{\oper} : ((\tau \to B) \times (\gamma \to (\later B))) \to B. \]
%
% This view makes it clear that $\sigma$ is the request type, i.e., the type of
% value sent to the environment, while $\tau$ is the response type, i.e., the type
% of value returned from the environment.

% We will often write B to denote either a $\Sigma$-algebra or its underlying
% set.

There is a forgetful functor $U_\Sigma$ from the category of $\Sigma-$algebras
and morphisms to the category of sets. The left adjoint to this functor takes a
set $A$ to the \textbf{free $\Sigma$-algebra} $F_\Sigma(A)$ on $A$. Explicitly,
the construction is by induction and guarded recursion. We seek to solve the
equation
%
\[ F_\Sigma(A) \cong A + \Sigma(F_\Sigma(A)), \]
%
i.e.,
%
\[ F_\Sigma(A) \cong A + \Sigma_{\oper: \Op}, 
    (\tau_\oper \to F_\Sigma(A)) \times G_\oper(\later F_\Sigma(A)). \]
%
First, consider the parameterized inductive type 
%
\[ P(X) := \mu Y. (A + \Sigma_{\oper : \Op}, (\tau_\oper \to Y) \times G_\oper(X)). \]
%
Now, using guarded recursion, define $F_\Sigma(A)$ to be the unique solution to
the guarded-recursive equation
%
\[ F_\Sigma(A) \cong P(\later F_\Sigma(A)). \]
%
Then by the definitions of least fixpoint and guarded fixpoint, we have that
$F_\Sigma(A)$ is a solution to the original equation.

Elements of $F_\Sigma(A)$ can intuitively be thought of as $\Sigma$-terms
generated by the elements of $A$ and the operations in $\Sigma$. We can view
such terms as a generalization of inductively-defined trees: a term can either
be a leaf, consisting of a variable in $A$, or a node labeled by an operation
$\oper : (\tau, \gamma)$, with $\tau$-many children available $\emph{now}$ and
$\gamma$-many children available $\emph{later}$.

% A similar notion of guarded-recursive trees has been studied under the name
% \emph{guarded interaction trees} \cite{guarded-interaction-trees}.

The type $F_\Sigma(A)$ has the universal property that given any
$\Sigma$-algebra $B$ and any map of variables $f : A \to B$, there is a unique
map $F_\Sigma(A) \to B$ extending $f$. In other words, if we specify how to
interpret the variables $A$ as elements of $B$, this extends uniquely to an
interpretation of any term. We write $\sem{t}$ to denote such an interpretation.
%
% Given any equation $e = (\Gamma_e, l_e, r_e)$, we can interpret $e$ in any
% $\Sigma$-algebra $B$ by asking that each side maps to the same term in $B$. This
% will be useful in defining the notion of guarded algebraic theory.

We now consider the addition of equations to a signature.
% A \textbf{guarded algebraic theory} $\calA$ consists of a guarded signature
% $\Sigma_\calA$ and a collection of equations $E_\calA$. 
A \textbf{guarded algebraic theory} $\calA$ is given by a signature
$\Sigma_\calA$ and a collection of equations $E_\calA$. An equation
$e$ consists of a context $\Gamma_e$ of free variables, and a pair of terms
$l$ and $r$ in $F_{\Sigma_\calA}(\Gamma_e)$.
%
If the signature $\Sigma_\calA$ is finitary, then we say that $\calA$ is a
\textbf{finitary theory}.

Models of a guarded algebraic theory are simply algebras of the associated
signature (i.e., a set $B$ with an interpretation of the operations in
$\Sigma_A$), in which all of the equations of the theory hold. More concretely,
a \textbf{model of a theory} $\calA = (\Sigma, E)$ is a model $B$ of the
signature $\Sigma$ such that for each equation $e = (\Gamma, l , r)$ in $E$, we have
$\sem{l} = \sem{r}$ for all instantiations $f : \Gamma_e \to B$ of the free
variables of $e$.
% i.e., for all $f : \Gamma_e \to B$.


% A \emph{model} of an algebraic theory $\calA$ is a set $B$ with an interpretation
% of the operations in $\Sigma_A$, such that all equations in $E_A$ hold. Formally
% speaking, this means that when we interpret the LHS and RHS of each equation as
% elements of $B$, those elements are equal.

For a guarded algebraic theory $\calA$, the category of $\calA$-models has a
forgetful functor $U_\calA$ to the category of sets, which simply takes the
underlying set of the model. This functor has a left adjoint $F_\calA$, which
given a set $X$ constructs the \textbf{free model} of $\calA$ on $X$. Concretely,
the free model of $\calA = (\Sigma_\calA, E_\calA)$ on a set $X$ is given by
quotienting the set of $\Sigma_\calA$-terms by the equations in $E_\calA$.
%
We can construct the free model in an analogous manner to our construction of
the free $\Sigma$-algebra for a signature $\Sigma$, using quotient-inductive
types in place of ordinary induction.
%
Under this construction, elements of $F_\calA(X)$ are mixed
inductive/guarded-recursive trees quotiented by the equations of the theory.
%
This adjunction induces a monad $T_\calA := U_\calA \circ F_\calA$ on the
category of sets, which is known as the \textbf{free model monad}.
% TODO cite
% For a given theory $\calA$, we denote the free model monad for $\calA$ by
% $T_\calA$.


% Because we are modeling gradual typing, at a minimum we want to ensure that our
% theory contains an operation corresponding to error, and since our semantics is
% intensional, i.e., step-aware, we also include an operation corresponding to
% taking an observable step. Such a step will occur whenever a non-well-founded
% recursive definition is unfolded; in the case of gradual typing this would be
% when inspecting a value of the dynamic type.
% % TODO cite GTT-SGDT paper?

\subsection{Examples of Algebraic Theories}

We introduce some specific guarded algebraic theories that will be used
throughout the paper.
%
\begin{example}[Exceptions]\label{ex:exn} Fix a type $E$. The theory of
    computations that can raise exceptions in $E$, denoted $\exn{E}$, is defined
    as follows. For each $e : E$, the signature $\Sigma_{\exn{E}}$ contains an
    operation $\ex{e}$ with $\tau_{\ex{e}} = 0$ and $\gamma_{\ex{e}} = 0$. There
    are no equations associated with this theory.

    Then a model of this theory is a set $B$ along with, for each $e : E$, a
    function $\sem{\ex{e}} : ((0 \to B) \times (0 \to \later B)) \to B$, i.e., a
    function $1 \to B$, which is just an element of $B$. In other words, models
    of this theory are simply pairs $(B, \{ \mho_e \}_{e : E})$ where $\mho_e
    \in B$ for each $e : E$.
  %
    The free-model monad for $\exn{E}$ is given by adjoining $E$ to its
    argument:
    %
    \[ F_{\exn{E}}(X) \cong X + E. \]
    
\end{example}
\begin{example}[Error]\label{ex:err}
    For the special case $\exn{\text{Unit}}$, i.e., where $E = \text{Unit}$, we
    call the resulting theory $\error$. Models of this theory are simply pointed
    sets $(B, \mho)$ where $\mho \in B$. The free-model monad for $\error$ is
    given by freely adjoining a point to its argument.
\end{example}

\begin{example}[Reader]\label{ex:reader} The theory of immutable state
  $\reader{R}$ models computations that can read from a single, global
  immutable variable of type $R$. The theory consists of a signature with a
  single operation $\lookup$ of arity $\tau_\lookup = R$ and guarded arity
  $\gamma_\lookup = 0$. In any algebra $B$ of this signature, the operation
  $\lookup$ has type $(R \to B) \to B$.
  %
  The lookup operation satisfies the following two equations:
  %
  \[ \forall (x : B).\, \lookup(\lambda r.\, x) = x, \]
  and
  \[ \forall (f : R \to R \to X).\,
    \lookup(\lambda r. \lookup (g\, r)) = \lookup(\lambda r.\, g\, r\, r\,). \]
  The free-model monad for this theory is the \emph{reader monad}, given by
  %
  \[ F_{\reader{R}}(X) \cong R \to X. \]
   
\end{example}

\begin{example}[Writer]\label{ex:writer} Fix a monoid $W$. The theory
  $\writer{W}$ models computations that can append elements in $W$ to an output.
  The theory consists of a signature with a family of operations $\wrt{w}$ for
  each $w \in W$ each of arity $\tau_{\wrt{w}} = 1$ and guarded arity
  $\gamma_{\wrt{w}} = 0$. In any algebra $B$ of this signature, the operation
  $\wrt{w}$ has type $B \to B$, with the argument representing what to do after
  appending $w$ to the output.
  %
  The operations $\wrt{w}$ satisfy the following equations. First, for all $x :
  B$, $\wrt{e_W}(x) = x$, where $e_W$ is the identity of $W$. This says that
  appending the identity element has no effect. Second, for all $w_1, w_2 : W$
  and $x : B$, we have $\wrt{w_1}(\wrt{w_2}(x)) = \wrt{w_1 \cdot w_2}(x)$, where
  $\cdot$ is the multiplication in $W$.

  Models of this theory are simply $W$-sets, i.e., sets equipped with an action
  of $W$. The free-model monad is the \emph{writer monad}, given by
  %
  \[ F_{\writer{W}}(X) \cong W \times X. \]
\end{example}

\begin{example}[State]\label{ex:state} %The theory of stateful computations
    The theory of global state ${\state{S}}$ for state set $S$ models
    computations that can read from and write to a global state --- a single
    global variable of type $S$. The theory consists of a signature
    $\Sigma_{\state{S}}$ with the following operations:
    %
    \begin{itemize}
        \item An operation $\get$ of arity $\tau_\get = S$ and guarded arity
        $\gamma_{\get} = 0$.
        \item A family of operations $\put_s$ for each $s \in S$, each of arity
        $\tau_{\put_s} = 1$ and guarded arity $\gamma_{\put_s} = 0$.
    \end{itemize}
    
    In any algebra $B$ of this signature, the operation $\get$ has type $(S \to
    B) \to B$ and $\put_s$ has type $B \to B$. The argument to $\put_s$
    represents the continuation, i.e., what we do after writing $s$ to the
    state.
    %
    In the theory of state, these operations are subject to the following four
    equations:
    %
    \begin{itemize}
        \item \textbf{get-get}: $\forall f : (S \to (S \to B)).$
            $\get(\lambda s. \get (f\, s)) = \get(\lambda s. f\, s\, s)$
        \item \textbf{get-put}:
            $\get(\lambda s.\put_s(x)) = x$
        \item \textbf{put-get}: $\forall g : (S \to B).\forall s : S.$
            $\put_s(\get g) = \put_s(g\, s)$.    
        \item \textbf{put-put}: $\forall x : B. \forall s, t : S.$
            $\put_t(\put_s(x)) = \put_s(x)$.
    \end{itemize}
    %
    Models of this theory are computations that can read from and write to a
    global state $S$.
    %
    The free-model monad for this theory is the \emph{global state monad}
    %
    \[ F_{\state{S}}(X) \cong S \Rightarrow (S \times X). \]

\end{example}

\begin{example}[Nondeterminism]\label{ex:nondeterminism}
  The theory of nondeterminism $\nondet$ models computations that can make
  nondeterministic choices. The signature $\Sigma_{\nondet}$ contains a single
  operation $\choose$ with $\tau_{\choose} = 2$ and $\gamma_{\choose} = 0$. The
  equations specify that the operation is associative, commutative, and
  idempotent.

  Models of this theory are join-semilattices, that is, a set $B$ equipped with
  a binary operation that is associative, commutative, and idempotent. The
  free-model monad for this theory is given by the finite powerset monad
  $P_f(X)$, the collection of all non-empty finite subsets of $X$; this is the
  free join-semilattice on $X$. It is possible to define $P_f(X)$ as a
  higher-inductive type \cite{10.1145/3167085, 10.3217/jucs-023-01-0063}.
\end{example}

In our next example, the guarded arity is nontrivial:
%
\begin{example}[Computational Steps]\label{ex:steps}
    %We define the theory $\calA_{step}$ of computational steps as follows. 
    The theory of computational steps, denoted $\steptheory$, models computations that
    can take an observable computational step. The theory is defined as follows.
    Let $\Sigma_{\steptheory}$ contain a single operation $\step$ with $\tau_{\step} =
    0$ and $\gamma_{\step} = 1$. There are no equations associated with this
    theory.

    Then a model of this theory is a set $B$ equipped with a function
    $\sem{\step} : ((0 \to B) \times (\later B)) \to B$, i.e., a function
    $\later B \to B$. We call this function $\theta_B$ in accordance with prior
    work using guarded type theory.
    %
    The free-model monad for $\steptheory$ is the so-called \emph{guarded-lift
    monad} \cite{mogelberg-paviotti2016} defined as the unique solution to the
    guarded-recursive equation
    %
    \[ L(X) \cong X + \later L(X). \]
    % TODO cite paper on guarded lift monad
\end{example}

%\max{state, reader, and non-determinism? Need to show off the generality here}

\subsection{Operations on Algebraic Theories}

We define two operations for combining algebraic theories, straightforward
generalizations of the sum and tensor of algebraic theories described by Hyland
et al. \cite{HYLAND200670}.

Given two guarded algebraic theories $\calA_1$ and $\calA_2$, we define their
\emph{sum} $\calA_1 + \calA_2$ by combining the operations and the equations in
$\calA_1$ and $\calA_2$. We do not impose any additional equations.
% TODO cite paper on combining algebraic theories.
% TODO example

Likewise, given two theories $\calA_1$ and $\calA_2$, we define their
\emph{tensor product} $\calA_1 \otimes \calA_2$ by combining the operations and
equations as in the coproduct, but also imposing additional equations that state
the commutativity of each of the operations in $\calA_1$ with all of those in
$\calA_2$.

\begin{example}
  The theory $\state{S} \otimes \steptheory$ models computations that can
  take observable computational steps, as well as read and write to a single
  global cell of type $S$. In addition to the equations from each theory (in
  this case, the state equations), the $\get$ and $\put_s$ operations must
  commute with the $\theta$ operation. This theory is described concretely by
  Møgelberg and Zwart \cite{lmcs:14660}, who showed that the free-model monad is
  given by $F(X) = S \to (L(S \times X))$, where $L$ is the guarded lift monad
  discussed above.
\end{example}

\begin{remark}\label{remark:tensor-nullary-with-step} Consider a theory $\calA$
  with a nullary operation $\oper$, i.e., $\tau_\oper = 0 = \gamma_\oper$. The
  tensor of $\calA$ with the theory of stepping $\calA_{\step}$ includes an
  equation $\theta(\lambda t.\, \oper) = \oper$. We can prove using
  \lob-induction that this implies $\oper = \fix\, \theta$, i.e., we cannot
  distinguish between performing $\oper$ and stepping indefinitely. As a result,
  we will avoid using the tensor to combine a theory containing a nullary
  operation with the theory of stepping.
\end{remark}

\subsection{Theory Transformers and Monad Transformers}

% Another operation on guarded algebraic theories involves augmenting a given
% theory $\calA$ with the operations and equations of a second theory $\calA_0$;
% we call such an operation a \textbf{theory transformer} for $\calA_0$.

Taking the sum or tensor with a fixed guarded algebraic theory gives a notion of
\textbf{theory transformer}.
%
For example, there is a theory transformer $M_{\exn{E}}$ for adding exceptions
to a guarded algebraic theory, given by $M_{\exn{E}}(\calA) := \calA + \exn{E}$.
Some theory transformers induce a corresponding \textbf{monad transformer} that
takes the free models of $\calA$ to the free models of the transformed theory.
For example, we have the following result for the global state transformer:
%
\begin{theorem}[State Monad Transformer]\label{thm:state-monad-transformer}  
Let $\calA$ be a guarded algebraic theory. If $T_\calA$ is the free-model monad
for $\calA$, then the free-model monad for the theory $\calA \otimes \state{S}$
is given by
%
\[ T'(A) := S \to T_\calA(S \times A). \]
%
\end{theorem}
%
Below we list the theory transformers we support, and their corresponding monad
transformers:
%
\begin{table}[h!]
  \centering
  \begin{tabular}{|c|c|c|}
    \hline
    Theory       & Theory Transformer         & Monad Transformer \\
    \hline
    Exceptions   & $- + \exn{E}$          & $T'(A) = T(A + E)$ \\
    Reader       & $- \otimes \reader{R}$ & $T'(A) = R \to T(A)$ \\
    Writer       & $- \otimes \writer{W}$ & $T'(A) = T(W \times A)$ \\
    Global State & $- \otimes \state{S}$  & $T'(A) = S \to T(S \times A)$ \\
    \hline
  \end{tabular}
  \label{tab:theory-and-monad-transformers}
\end{table}

% \max{Shouldn't this be the part where we discuss monad transformers as
%   well? I think this is where we should basically list theory/monad
%   transformers we handle and what you can get out of them.}

\subsection{Relational Aspects}

\max{This paragraph could be more concise and more informative}
An important aspect that we have elided so far is the fact that, since we are
modeling gradual typing, we need our mathematical structures to be equipped with
a notion of relation to model the graduality ordering. This theory was developed
extensively by Giovannini et al. in their denotational model for gradual typing
\cite{gtt-sgdt}. The implication of this is that the categories involved in the
adjunction described above are not merely the category of sets and the category
of $\calA$-models. We require additional relational structure: a relation to
model the graduality ordering as well as a notion of weak bisimilarity that
relates terms that differ only in the number of steps they take. These relations
are crucial ingredient in the graduality proof. Since these relational aspects
are not relevant for the operational semantics, we defer a detailed discussion
to the section on denotational semantics.

% for now, it is sufficient to keep in mind that the categories involved, and the
% free-model monad $T_\calA$ for a given algebraic theory $\calA$, enjoy this
% additional structure.


\section{A Generic Syntax for Gradual Typing with Effects}\label{sec:syntax}

To begin, we fix a framework for specifying gradual cast calculi with algebraic
effects. The framework is parameterized by an ordinary \emph{finitary} algebraic
theory $\calA$ specifying the effects available in the syntax. Finitary here
means that the arities of all operations are finite types; we write $[n]$ for
the finite type with $n$ elements, unique up to isomorphism.

The cast calculus is call-by-value and is presented in Figure
\ref{fig:gtlc-syntax}.

% TODO should we use CBPV syntax instead?
% TODO what about coproducts and lazy products?
% TODO if we are doing fine-grain CBV, do we need return and bind?

Types include natural numbers, functions $A \ra A'$, products $A \times A'$, and
a dynamic type, which is the type assigned to a dynamically-typed value, i.e., a
value whose type is unknown at compile-time.

As is typical in cast calculi, there is an ordering on types known as \emph{type
precision} specifying when one type is more precise (i.e., less dynamic) than
another. Following prior work, our calculus uses an explicit syntax of type
precision \emph{derivations} of the form $c : A \ltdyn A'$. Such a derivation
should be thought of as a proof that $A \ltdyn A'$. This presentation is
particularly convenient for carrying out proofs by induction.
%
As in prior work, the rules for precision include reflexivity $r(A)$,
transitivity $cc'$, and monotonicity $c \ra c'$ and $c \times c'$, as well as
generators $\inat,\iarr,\itimes$ that correspond to the tags of the dynamic
type.

Following Giovannini et al., we also specify an equational theory $c \equiv c'$
on the type precision derivations; the idea is that that while the denoted
relations are not necesseasaily equal in the denotational semantics, they are
\emph{quasi-equivalent}: if two terms are related by $\sem{c}$, then they are
related by $\sem{c'}$ other up to the insertion of delays, and vice versa. See
their paper for more details \cite{gtt-sgdt}.

The cast calculus has special coercion terms called casts, which come in two
forms: upcasts and downcasts. Given $c : A \ltdyn A'$ and $M$ of type $A$, the
term $\upc c M$ has type $A'$, and given $N$ of type $A'$, the term $\dnc c N$
has type $A$. Upcasts are pure, i.e., they always succeed (at least to a first
order), whereas downcasts are effectful.

% Lastly, we have terms $\synop{op}{M}$ corresponding to each operation in the
% algebraic theory $\calA$. More concretely, if $\eff{op}{\sigma}{\tau}$ is an
% operation in $\calA$ and $M$ is of type $\sigma$, then $\synop{op}{M}$ is of
% type $\tau$.

Lastly, we have terms $\synoper{M}{n}$ corresponding to each operation in the
algebraic theory $\calA$. More concretely, if $\oper$ is an operation in $\calA$
of finite arity $n$, then given $n$ terms of type $A$ (representing the $n$
possible continuations), the operation $\synoper{M}{n}$ has type $A$.

% We mention two notes about the relationship between the given algebraic theory
% and the syntax. First, 

% We assume that the meta-level arity types of each operation exist as
% object-level types within the cast calculus. While these types are not
% \emph{equal}, there will be an isomorphism allowing us to translate between
% them. For example, if we have an operation $\oper$ of arity $\mathbb{N}$, then
% in the syntax the corresponding term will have type $\nat$. We leave implicit
% the conversion between the syntactic and semantic versions of the arities.


% Terms:
% upcasts and downcasts
% effect operations

% explain type precision

% New, Licata, and Ahmed gave a syntax and logic axiomatizing the graduality
% property \cite{new-licata-ahmed2019}.

\begin{figure}
  \begin{mathpar}
    \begin{array}{rcl}
    \text{Types } A &::=
      &\nat 
      \alt \,\dyn 
      \alt A \ra A' 
      \alt A \times A'\\
    \text{Type Precision } c &::=
      & r(A) 
      \alt c c'     \alt \iarr 
      \alt \inat    \alt \itimes 
      \alt c \ra c' \alt c \times c'\\
    \text{Values } V &::=
      & x 
      \alt \upc \inat V \alt \upc \itimes V \alt \upc \iarr V \alt \zro 
      \alt \suc\, V \alt \lda{x}{M} 
      \alt (V,V') \\ 
    \text{Terms } M,N &::=
      & \err 
      \alt \upc c M  \alt \dnc c M 
      \alt \zro      \alt \suc\, M 
      \alt \lda{x}{M} \\ 
      &&\alt M\, N   \alt (M,N) 
      \alt \textrm{split } (x,y) = M \textrm{ in } N 
      %\alt \textrm{let } x = M \textrm{ in } N 
      \alt \synop{op}{V}{x}{M} \\
    \text{Contexts } \Gamma &::= &\cdot \alt \Gamma, x : A \\
    \text{Ctx Precision } \Delta &::=& \cdot \alt \Delta,x:c \\
    \text{Evaluation Contexts } E &::= & \bullet
      \alt \textrm{split } (x,y) = E \textrm{ in } N
      \alt (E, M)
      \alt (V, E) \\
      %\alt \textrm{let } x = E \textrm{ in } N
      &&\alt \suc\, E
      \alt E\, M
      \alt V\, E
      \alt \upc c E 
      \alt \dnc c E
  \end{array}

  \inferrule*[right=Up]
  {\Gamma \vdash M : A \and c : A \ltdyn A'}
  {\Gamma \vdash \upc c M : A'}

  \inferrule*[right=Dn]
  {\Gamma \vdash N : A' \and c : A \ltdyn A'}
  {\Gamma \vdash \dnc c N : A}

  \inferrule*[right=Eff]
  {\oper : n 
    \and \forall i \in [n]. \Gamma \vdash M_i : A}
  {\Gamma \vdash \synoper{M}{n} : A}

  \inferrule{}{\Gamma \vdash \mho : A}
  \end{mathpar}
  \hrule
  \begin{mathpar}
    \inferrule{}{r(A) : A \ltdyn A}\and
    \inferrule{c : A \ltdyn A' \and c' : A' \ltdyn A''}{cc' : A \ltdyn A''}\and
    \inferrule{}{\iarr \colon \dyn \ra \dyn \ltdyn \dyn}\and
    \inferrule{}{\inat \colon \nat \ltdyn \dyn}\and
    \inferrule{}{\itimes \colon \dyn \times \dyn \ltdyn \dyn}\and
    \inferrule{c_i : A_i \ltdyn A_i' \and c_o : A_o \ltdyn A_o'}{c_i \ra c_o : (A_i \ra A_o) \ltdyn (A_i' \ra A_o')}\and
    \inferrule{c_1 : A_1 \ltdyn A_1' \and c_2 : A_2 \ltdyn A_2'}{c_1 \times c_2 : (A_1 \times A_2) \ltdyn (A_1' \times A_2')}\and
  \end{mathpar}
  \hrule
  \begin{mathpar}
     r(A)c \equiv c\and
     c \equiv cr(A')\and
     c(c'c'') \equiv (cc')c''\and
     r(A_i \ra A_o) \equiv r(A_i) \ra r(A_o)\and
     r(A_1\times A_2) \equiv r(A_1) \times r(A_2)\and
     (c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra c_oc_o') \and
     (c_1\times c_2)(c_1'\times c_2')\equiv (c_1c_1' \times c_2c_2')
  \end{mathpar}
  \caption{Syntax for Gradual Typing with Effects}
  \label{fig:gtlc-syntax}
\end{figure}


% rules for term precision
Next, we have the axioms for reasoning about terms. First is the equational
theory, including the usual CBV $\beta\eta$ rules for function and product
types.
%
We also have term precision, the extension of type precision to terms. Figure
\ref{fig:term-prec} presents selected rules for the term precision relation.
%
As in prior work, term precision rules have the form $\Delta \vdash M \ltdyn M'
: c$ where $\Delta$ is a context mapping variables to type precision
derivations.
%
% The judgment is only well formed when every use of $x : c$ for $c : A
% \ltdyn A'$ is used with type $A$ in $M$ and $A'$ in $M'$ and similarly
% the output types match $c$.

The term precision rules are as follows. First, the rule \textsc{ErrBot} says
that the error term $\mho$ is the most precise term.
%
Next, we have a congruence rule for each term constructor, e.g., the rule for
application says that if $M \ltdyn M'$ and $N \ltdyn N'$ then $M\,N \ltdyn
M'\,N'$. For the sake of brevity, the congruence rules are elided from the
figure.\max{TODO: add congruence for operations}
%
Using these congruence rules, reflexivity $M \ltdyn M$ is admissible. As in the
syntax of Giovannini et al. \cite{gtt-sgdt}, and in accordance with the original
formulation of the dynamic gradual guarantee \cite{siek_et_al:LIPIcs:2015:5031},
we do not take transitivity of term precision as a primitive rule.

As in prior work, we have a rule \textsc{EquivTyPrec} stating that equivalent
type precision derivations $c \equiv c'$ are interchangeable for the purposes of
term precision \cite{gtt-sgdt}.

Lastly, we have the four fundamental rules for reasoning about casts (\upl,
\upr, \dnl, \dnr), which axiomatize the upcasts and downcasts as behaving like
least upper bounds and greatest lower bounds, respectively. 

% these say that the upcast is a kind of \emph{least upper bound} and dually that
% the downcast is a \emph{greatest lower bound}.


\begin{figure}
  \begin{mathpar}

  \inferrule*[right=EquivTyPrec]
  {\Delta\vdash M \ltdyn M' : c \and c \equiv c'}
  {\Delta\vdash M \ltdyn M' : c'}

  \inferrule*[right=ErrBot]
  {}
  {\Delta \vdash \mho \ltdyn M : c}

  \inferrule*[right=UpL]
  {M \ltdyn M' : cc_r}
  {\upc {c} M \ltdyn M' : c_r}

  \inferrule*[right=UpR]
  {M \ltdyn M' : c_l}
  {M \ltdyn \upc {c} M' : c_lc}

  \inferrule*[right=DnL]
  {M \ltdyn M' : c_r}
  {\dnc {c} M \ltdyn M' : cc_r}

  \inferrule*[right=DnR]
  {M \ltdyn M' : c_lc}
  {M \ltdyn \dnc {c} M' : c_l}
  \end{mathpar}
  \caption{Term Precision Rules (Selected)}
  \label{fig:term-prec}
\end{figure}

Our goal in the remainder of this work is to define an operational semantics for
this cast calculus and prove that it satisfies graduality. For an outline of our
approach to doing so, refer to Section \ref{sec:overview-of-approach}.

% compositional denotational semantics of types, terms, type and term precision
% from which we can easily extract a big step semantics that satisfies graduality
% and respects the equational theory of the calculus.


\section{Operational Semantics}\label{sec:operational-semantics}

In this section, we define a small-step operational semantics for the generic
cast calculus introduced in Section \ref{sec:syntax}. Our approach is based
loosely on the recent work of Kavvos \cite{kavvos-2025}, who gave a generic
operational semantics for a language with algebraic effects.

Recall that our framework is parameterized by an \emph{ordinary} algebraic
theory $\calA$; in particular, this means there are no guarded operations in
$\calA$.
%
Because we are modeling gradual typing, at a minimum we want to ensure that our
theory contains an operation corresponding to error, and since our semantics is
intensional, i.e., step-aware, we also include an operation corresponding to
taking an observable step. Such a step will occur whenever a non-well-founded
recursive definition is unfolded; in the case of gradual typing this would be
when inspecting a value of the dynamic type.
%
In fact, for reasons that will become clear later in the present section, we
distinguish between two different kinds of semantic stepping operations:
\textbf{true steps}, which occur when inspecting a non-well-founded definition
such as the dynamic type, and \textbf{bureaucratic steps}, which represent all
other steps in the operational semantics. A key difference between the two is
that bureaucratic steps are invisible in the denotational semantics; they are
merely a bookkeeping tool necessary in defining the big-step operational
semantics.
%
If we view steps as having an associated ``cost'', we can think of true steps as
having a nonzero cost, while bureaucratic have zero cost.
%
% TODO: is this accurate?
The logical relation defined in Section \ref{sec:relating-the-semantics} will
imply that when running any term, there can be at most finitely many consecutive
bureaucratic steps between any two operations.

To account for errors and computational steps, the construction of the
operational semantics is parameterized by three guarded algebraic theories that
extend $\calA$. We write $\calA[\oper_1, \oper_2, \dots, \oper_n]$ to denote the
collection of guarded theories that have the operations and equations of $\calA$
plus the operations $\oper_1, \dots, \oper_n$, and possibly contain equations
relating $\oper_i$ to the operations of $\calA$. The guarded extensions taken as
input are as follows:
%
\begin{enumerate}
  \item $\calAerr \in \calA[\mho]$, i.e., $\calA$ extended with errors.

  \item $\calAs \in \calAerr[\theta^s]$, i.e., $\calA$ extended with errors and
  (true) steps.

  \item $\calAsb \in \calAs[\theta^b]$, i.e., $\calA$ with errors, true steps,
  and bureaucratic steps.
  % \max{define macros for this and make sb not be in textrm}
  % % 
  % $\calA_{\mho sb} := 
  %   (\calA \otimes (\th{\mho} + \th{\step} + \th{\step}))$. 
%
  % We denote the first step operation by $\theta^{s}$ and the second step
  % operation by $\theta^{b}$.
\end{enumerate}

We write $T_\mho$, $\Ts$ and $\Tsb$ to denote the free-model
monads for the theories $\calAerr$, $\calAs$, and $\calAsb$, respectively.
(Recall that the free-model monad for a theory is unique up to isomorphism of
monads.)
%
The monad $T_{\mho}$ supports each of the operations of the input theory
$\calA$, in addition to a nullary error operation $\mho$. This means that for
each operation $\oper$, of arity $\tau_\oper$, there is a function
$\sem{\oper}^{T X} : (\tau_\oper \to T X) \to T X$.
%
The monad $\Ts$ supports each of the above, in addition to a guarded step
operation $\theta^s : \later \Ts(X) \to \Ts(X)$.
%
The same is true of $T_{\mho sb}$, but it has two guarded step operations
$\theta^s, \theta^b : \later \Tsb(X) \to \Tsb(X)$.

% Since $\calA_{\mho}$ is a subtheory of $\calA_{\mho + s}$, there is a morphism
% of monads $\iota : T_\mho \to T_{\mho s}$ and likewise there is a morphism of
% monads $\iota' : T_{\mho s} \to T_{\mho sb}$.

% TODO: are specific presentations of the monads taken as input?

\subsection{Defining the Operational Semantics}

% TODO: define the theory of \calA + stepping

In what follows, we use $\term{A}$ to denote the (meta-theoretic) type of closed
terms of type $A$ in the cast calculus, and likewise $\val{A}$ denotes the type
of closed values of type $A$ in the cast calculus.

Our goal is to define a guarded big-step operational semantics, i.e., a function
%
\[ \run : \term{A} \to T_{\mho sb} (\val{A}). \]
%
Observe that the output of running a term is a monadic value. The monad accounts
for all possible effects that may occur as the term is being run, including both
true and bureaucratic steps.
%
We write $M \Dwn \hat{V}$ to mean that $\run(M) = \hat{V}$ and we write $M
\Uparrow$ to mean that $\run(M) = \fix \thetatrue$.


To define the $\run$ function, we first model small-step reduction as a function
%
\[ \step : (M : \term{A}) \to (\Sigma_{V \in \val{A}} (M = V)) 
    + (\term{A} + \Sigma_\mho(\term{A}) \times \steptype). \]
%
Here, $\Sigma_\mho$ is the signature of the algebraic theory $\calA_\mho$, and
$\steptype$ is simply a type with two inhabitants: $\steptype := \tstep \alt
\bstep$, where $\tstep$ represents a true step and $\bstep$ represents a
bureaucratic step.
%
Then the $\step$ function states that for every closed term $M$ of type $A$,
either $M$ is a value or it takes a step. When a term steps, it can be a
\textbf{pure reduction step} to some other term $M'$, or an \textbf{effectful
reduction step}, involving an operation in the input theory $\calA$ or error.
For both kinds of reduction, we also track the nature of the step (true or
bureaucratic).

% Notation for stepping
% For the sake of brevity, we use the following notation for stepping. If
% $\step{M} = \inr(\hat{N}, ty)$, we say that $M \stepsin{0} \hat{N}$ if $ty =
% \bstep$ and $M \stepsin{1} \hat{N}$ if $ty = \tstep$. That is, the natural
% number (either $0$ or $1$) indicates whether the step is a true step or an
% bureaucratic step. In the cost analogy discussed above, the natural number
% represents the ``cost'' of performing the step, whereby true steps have cost $1$
% and bureaucratic steps have cost $0$.
% %
% For a pure reduction step $\hat{N} = \inl N$, we omit the $\inl$ and simply
% write $M \stepsin{k} N$.

We define an endofunctor $\Sigmahat(X) = X + \Sigma_\mho(X)$. Its action on a
function $f : X \to Y$ is defined in the obvious way, using $f$ in the first
case, and the action of $\Sigma_\mho$ on $f$ in the second case.
%
We write $\cpure{M}$ for the left injection into $\Sigmahat(X)$, and
$\ceff{\bar{M}}$ for the right injection.

For the sake of brevity, we use the following notation for stepping. We write $M
\stepsin{k} \hat{N}$, where $\hat{N}$ is a metatheoretic value of type
$\Sigmahat(\term{A})$, to mean that $\step(M) = \inr (\hat{N}, t)$ where $t =
\tstep$ if $k = 1$ and $t = \bstep$ if $k = 0$. That is, the number $k$ (either
$0$ or $1$) indicates whether the step is a true step or a bureaucratic step. In
the cost analogy discussed above, $k$ represents the ``cost'' of performing the
step, where true steps have cost $1$ and bureaucratic steps have cost $0$.
%
For a pure reduction step $\hat{N} = \cpure N$, we often omit the $\cpure{}$ and
simply write $M \stepsin{k} N$.



% Old version:
% To define the $\run$ function, we first model small-step reduction as a function
% %
% \[ \step : (M : \term{A}) \to (\Sigma_{V \in \val{A}} (M = V)) + (T(\term{A}) \times \steptype). \]
% %
% That is, given a term of type $A$, either that term is a value, or it takes an
% effectful step. Here, $\steptype$ is simply a type with two inhabitants:
% $\steptype := \tstep \alt \bstep$, where $\tstep$ represents a true step and
% $\bstep$ represents a bureaucratic step.
% %
% Notice that the small-step function, by virtue of its type signature, includes
% the usual progress, preservation, and determinacy theorems.



% Goal: operational graduality
% TODO: should we move this to before the operational semantics section?
% Our main goal in the rest of the paper will be to prove that this big-step
% semantics validates the graduality property:

% \begin{theorem}[Graduality]
%   Suppose $M \ltdyn N : \nat$. Then either:
%   \begin{enumerate}
%     \item $M \Dwn \mho$, where $\mho \in T_{db} (\val{\nat})$ is the error element of the monad.
%     \item $M \Dwn \eta\, n$ and $N \Dwn \eta\, n$ where $n \in \nat$.
%     \item $M \Uparrow$ and $N \Uparrow$.
%     \item $M \Dwn \text{op}(V)$ and $N \Dwn \text{op}(V)$ % M and N both terminate with a 0-ary effect
%   \end{enumerate}
% \end{theorem}

\subsection{Defining the Small-Step Semantics}\label{sec:small-step-semantics}

Figure \ref{fig:operational-semantics} defines the small-step semantics. The
presentation is as a series of rules, but these can be converted to the form of
the $\step$ function whose signature we have discussed above:
%
\[ \step : (M : \term{A}) \to (\Sigma_{V \in \val{A}} (M = V)) 
    + (\term{A} + \Sigma_\mho(\term{A}) \times \steptype). \]
%   
In particular, our notational conventions are as follows:
%
\begin{itemize}
  \item A reduction rule $M \stepsin{k} \hat{N}$, where $\hat{N}$ is of type
  $(\term{A} + \Sigma_\mho(\term{A})) \times \steptype)$, should be translated as
  $\step(M) = \inr (\hat{N}, t)$ where $t = \tstep$ if $k = 1$ and $t = \bstep$
  if $k = 0$. That is, the number $k$ (either $0$ or $1$) determines whether the
  step is considered a true step or a bureaucratic step.

  \item A \emph{pure reduction rule} is a special case of the above where
  $\hat{N} = \cpure{N}$ for some term $N$. In this case, we write $M
  \stepsin{k} N$, where $N$ is a term, leaving the $\cpure{}$ constructor
  implicit.

  \item An \emph{impure reduction rule} is one where $\hat{N} = \ceff{(\oper,
  u)}$ for some $\oper : \calA_\mho.\Op$ and $u : \tau_\oper \to \term{A}$. Note
  that every operation in $\calA_\mho$ has trivial guarded arity, so there are
  no guarded values.
  
  \item Any term $M$ that does not match one of these rules is equal to a value
  $V$, and hence $\step(M)$ returns $\inl$ applied to the value $V$ and the
  proof that $M$ is equal to $V$.

  \item We write $\mho$ to denote the element $(\err, *)$ of
  $\Sigma_\mho(\term{A})$.
\end{itemize}

% TODO cite wrapping semantics for casts
% TODO define/explain tags
% TODO make sure to distinguish between meta-level lambda and object-level lambda
% TODO fix the congruence rules for evaluation contexts

\begin{figure}

  \textbf{Beta rules:}

  \begin{mathpar}

    % Usual beta rules:
    \inferrule*[right=Beta]
      { }
      {(\lambda x : A.M)(V) \stepsin{0} M[V/x]}

    \inferrule*[right=Split]
      { }
      {\synsplit{(V_1, V_2)}{x_1}{x_2}{M} \stepsin{0} M[V_1, V_2 / x_1, x_2]}

    % \inferrule*[right=Bind]
    %   { }
    %   {\synbind{\ret\, V}{x}{M} \stepsin{0} M[V/x]}
  \end{mathpar}

  \textbf{Casts (functorial actions and composition):}
  \begin{mathpar}

    % Upcast of a function steps when applied to a value:
    \inferrule*[right=Up-Arr]
      { c_i : A_i \ltdyn A_i' \and c_o : A_o \ltdyn A_o' 
        \and \Gamma \vdash V_f : A_i \ra A_o 
        \and \Gamma \vdash V : A_i'}
      { (\upc (c_i \ra c_o) V_f) \stepsin{0} (\lambda x. (\upc {c_o} (V_f\, (\dnc {c_i} x)))) }

    % Downcast of a function steps when applied to a value:
    \inferrule*[right=Dn-Arr]
      { c_i : A_i \ltdyn A_i' \and c_o : A_o \ltdyn A_o' 
        \and \Gamma \vdash V_f : A_i \ra A_o 
        \and \Gamma \vdash V : A_i'}
      { (\dnc {(c_i \ra c_o)} V_f) \stepsin{0} (\lambda x. (\dnc {c_o} (V_f\, (\upc {c_i} x)))) }


    % Up/down cast of a paired value steps to the pair of the up/down casted
    % values:
    \inferrule*[right=Up-$\times$, leftskip=1em]
      { c_1 : A_1 \ltdyn A_1' \and c_2 : A_2 \ltdyn A_2' 
        \\\\ \Gamma \vdash V_1 : A_1 \and \Gamma \vdash V_2 : A_2}
      { \upc {(c_1 \times c_2)} (V_1, V_2) \stepsin{0} (\upc {c_1} V_1, \upc {c_2} V_2) }

    \inferrule*[right=Dn-$\times$, leftskip=2em]
      { c_1 : A_1 \ltdyn A_1' \and c_2 : A_2 \ltdyn A_2' 
        \\\\ \Gamma \vdash V_1 : A_1' \and \Gamma \vdash V_2 : A_2'}
      { \dnc {(c_1 \times c_2)} (V_1, V_2) \stepsin{0} (\dnc {c_1} V_1, \dnc {c_2} V_2) }
  

    % "Composition" of casts with inj-arr and inj-times:
    \inferrule*[right=CmpUp]
      { c : A_1 \ltdyn A_2 \and c' : A_2 \ltdyn A_3 \\\\ \Gamma \vdash V : A_1 }
      { \upc {(cc')} V \stepsin{0} (\upc c' (\upc c V)) }

    \inferrule*[right=CmpDn]
      { c : A_1 \ltdyn A_2 \and c' : A_2 \ltdyn A_3 \\\\ \Gamma \vdash V : A_3 }
      { \dnc {(cc')} V \stepsin{0} (\dnc c (\dnc c' V)) }

    % \inferrule*[right=$\iarr$Up, leftskip=1em]
    %   { c : A \ltdyn (\dyntodyn) \and \Gamma \vdash M : A }
    %   { (\upc (\injarr c) M) \stepsin{0} (\upc \iarr (\upc c M))}

    % \inferrule*[right=$\iarr$Dn, leftskip=2em]
    %   { c : A \ltdyn (\dyntodyn) \and \Gamma \vdash M : \dyn }
    %   { (\dnc (\injarr c) M) \stepsin{0} (\dnc c (\dnc \iarr M))}

    % \inferrule*[right=$\itimes$Up]
    %   { c : A \ltdyn (\dyn \times \dyn) \and \Gamma \vdash M : A }
    %   { (\upc (\injtimes c) M) \stepsin{0} (\upc \itimes (\upc c M))}

    % \inferrule*[right=$\itimes$Dn, leftskip=2em]
    %   { c : A \ltdyn (\dyn \times \dyn) \and \Gamma \vdash M : \dyn }
    %   { (\dnc (\injtimes c) M) \stepsin{0} (\dnc c (\dnc \itimes M))}

  \end{mathpar}

  \textbf{Casts involving dyn:}
  \begin{mathpar}
    % Downcasts of upcasted-to-dyn values:
    \inferrule*[right=DnUpInjNat]
      { \Gamma \vdash V : \nat }
      { \dnc \inat (\upc \inat V) \stepsin{0} V}

    \inferrule*[right=DnUpInj$\times$]
      { \Gamma \vdash V : \dyn \times \dyn }
      { \dnc {\itimes} (\upc {\itimes} V) \stepsin{0} V}

    \inferrule*[right=DnUpInj$\arr$, leftskip=1em]
      { \Gamma \vdash V : \dyntodyn }
      { \dnc {\iarr} (\upc {\iarr} V) \stepsin{1} V}

    \inferrule*[right=TagMismatch, leftskip=1em]
     { \text{tag}_1, \text{tag}_2 \in \{ \mathbb{N}, \times, \ra \} \and \text{tag}_1 \neq \text{tag}_2 }
     { \dnc {\text{Inj}_{\text{tag}_1}} (\upc {\text{Inj}_{\text{tag}_2}} V) \stepsin{0} \ceff{(\mho, *)} }


  \end{mathpar}

  \textbf{Effects and congruence rules:}
  \begin{mathpar}

    % Effects
    % \inferrule*[right=Eff]
    %   { \eff{op} \sigma \tau }
    %   { \synop{op} V x M \stepslabel{\eff {op} V W} M[W/x] }

    \inferrule*[right=Eff, leftskip=2em]
      { \oper : [n] }
      { \synoper{M}{n} \stepsin{0} \ceff{(\oper, (\lambda\, (i : [n]).\, M_i))} }  

    % Congruence
    % Accounts for the fact that reduction may be impure
    \inferrule*[right=Cong]
      { i \in \{0,1\} \and M \stepsin{i} \hat{N}}
      { E[M] \stepsin{i} \Sigmahat(E[\cdot])(\hat{N}) }

    % \inferrule*[right=EffCong]
    %   {M \stepslabel {\eff {op} V W} N}
    %   {E[M] \stepslabel {\eff {op} V W} E[N]}


  \end{mathpar}
  \caption{Operational Semantics (Selected Rules)}
  \label{fig:operational-semantics}
\end{figure}

There are $\beta$ rules for functions, pairs, and bind, which are standard. Next
are the rules involving casts. We first have rules involving upcasts and
downcasts of functions and products.
%
For function casts, we use the standard \emph{wrapping semantics} of contracts
\cite{findler-felleisen02}. In particular, when a function $V_f$ is upcasted, it
steps to a lambda expression. When the upcasted function is applied to a value,
the value is downcasted according to the domain and applied to the original
function $V_f$, and then the result is upcasted according to the codomain. The
dual holds for a function that is downcasted.
%
When a pair of values $(V_1, V_2)$ is upcasted, we simply upcast each component
of the pair, and likewise for downcasting.
%
Next we have rules for up/down casts involving composed precision derivations,
i.e., $\injarr c$ and $\injtimes c$. These rules decompose such casts into the
cast for $c$ and the primitive injection casts $\iarr$ or $\itimes$. 
% These rules express the behavior of such casts in
% terms of the behavior of the cast for $c$ and the behavior of the primitive
% injection casts $\iarr$ or $\itimes$. 
For example, to upcast a term using the precision derivation $\injarr c$ where
$c : A \ltdyn (\dyntodyn)$, we first upcast the term by $c$, and then upcast the
result by $\iarr : \dyntodyn \ltdyn \dyn$.
%
Finally, we have the rules involving downcasting a value that was cast to $\dyn$
using the injection casts $\inat$, $\injarr c$, and $\injtimes c$. In each case,
we check that the tag of the upcast matches the tag of the downcast, e.g., the
upcast is $\iarr$ and the downcast is $\iarr$. If the tags match, then the cast
steps to the original value. Otherwise, the term takes an effectful step to
$\ceff{(\mho, *)}$, representing that an error was raised. Note that when
downcasting a dynamically-typed function, we record a true step, i.e., a
recursive unfolding. In the denotational semantics, this corresponds to
inserting a $\theta$.
%
% Note also that these rules only mention the primitive injection forms $\inat$,
% $\iarr$, and $\itimes$ rather than the composed forms $\injarr c$ and $\injtimes
% c$. This is sufficient because casts involving $\injarr c$ or $\injtimes c$ will
% first step to terms involving $\iarr$ and $\itimes$, which can then be handled
% using the rules just discussed.

Next, we have our second impure reduction rule. This rule states that for an
operation $\oper$ with arity $n$, the term $\synoper{M}{n}$ takes an effectful
step to the metatheoretic value $\ceff{(\oper, (\lambda\, (i : [n]).\, M_i))}$.
The intuition is that we are performing the operation at the meta-level.
%
Note that the operation is uninterpreted at this stage, i.e., there is no
environment to provide the value $i$. Instead, when we interpret the semantics
in a monad supporting the effects, we will invoke the corresponding operation on
the monad. The continuation will then select the appropriate term $M_i$ based on
the value of $i$ it receives from the environment.
% Note that since $\tau$ is a metatheoretic type, strictly speaking a conversion
% is necessary to substitute $z$ in the object-level term $M$. As mentioned in
% Section \ref{sec:syntax}, we assume that the arity of each operation is
% isomorphic to a corresponding syntactic type. For the sake of simplicity, we
% elide these conversions.

% for any possible request value $V$ and
% response value $W$, the term $\synop{op} V x M$ steps to $M[W/x]$. This
% represents the fact that the environment has responded with value $W$, after
% which the program continues as $M$ with $W$ in place of $x$.

The final rule says that we can perform pure and impure reductions under an
evaluation context. Recall that $\Sigmahat$ is the endofunctor defined by
$\Sigmahat(X) = X + \Sigma_\mho(X)$. The function $E[\cdot]$ is shorthand for
the function taking a term $M$ to $E[M]$.
% The final rules are simply congruence rules allowing us to perform pure and
% impure reductions under an arbitrary evaluation context.

\subsection{Defining a Runnable Semantics}\label{sec:big-step-semantics}

Now that we have defined a small-step semantics, we discuss how to extract a
runnable ``big-step'' semantics. More concretely, recall that our goal is a function
%
\[ \run : \term{A} \to T_{\mho sb} (\val{A}). \]
%
Notice that the codomain allows for two kinds of steps: true steps, and
bureaucratic steps. It is necessary that the monad be able to accommodate both
kinds of steps. 

% On the one hand, we want to preserve the true steps taken by terms so that we
% can relate them to the recursive unfoldings on the denotational side. On the
% other hand, because it is not necessarily the case that there are an 

We define $\run$ by guarded recursion to be the unique function satisfying the
properties
%
\begin{mathpar}
  \inferrule*
    { \step{M} = \inl (V, eq) }
    { \run{M} = \Tsb.\eta(V) }

  \inferrule*
    { \step{M} = \inr (\hat{N}, \mathit{ty}) }
    { \run{M} = \delta_{\mathit{ty}} (\run^{\dag} {(f(\hat{N}))}) }
\end{mathpar}
%
In the above:
%
\begin{itemize}
  \item $f : (\term{A} + \Sigma_\mho(\term{A})) \to \Tsb(X)$ is defined by
  cases, mapping $\cpure{M}$ to $\eta^{\Tsb}(M)$, and mapping $\ceff{(\oper, u, g)}$ 
  to $\Tsb(\term{A}).\oper(\eta \circ u, (\later \eta) \circ g)$.
  (Note that $u : \tau_\oper \to \term{A}$ and $g : \gamma_\oper \to \later (\term A)$.)
  \item The variable $\mathit{ty}$ has type $\steptype$.
  \item $\delta_{\mathit{ty}} = \theta_{\mathit{ty}} \circ \nxt$ is the
  ``delay'' function on $\Tsb$. The type of delay is determined by
  $\mathit{ty}$.
\end{itemize}
%
Notice that the recursive call to $\run$ is guarded by a $\theta$, so the
guarded fixpoint operator provides a unique function satisfying the above
properties.

We next define a function that ``runs'' an evaluation context $E$. Specifically,
if $E : A \multimap B$, then we define $\run{E} : \Tsb(\term A) \to
\Tsb(\val B)$ as
%
\[ \run{E} := (\lambda M. \run{(E[M])})^\dag, \]
%
where $\dag$ is the Kleisli extension operator for the monad $\Tsb$.

Then we have the following lemma, which says that running commutes with filling
an evaluation context:
%
\begin{lemma}[Run is a Homomorphism]\label{lem:run-fill}
  $\run{(E[M])} = (\run{E})(\run{M})$.
\end{lemma}
% TODO cite appendix.
The proof uses \lob-induction; see the Appendix for details.


\max{I think we need examples here of how this is really the normal operational semantics when we take specific examples like state, reader, probability etc.}
\eric{I added an example with state.}

We conclude this section with an example of how our semantics works for a
specific algebraic theory. Consider the theory of global state as discussed in
Section \ref{sec:modeling-effects}, and suppose the type $S$ is taken to be
$\bool$. The free models for this theory augmented with errors, true delays, and
bureaucratic delays are given by $T(X) := \bool \to L(\bool \times X)$ where $LX
\cong X + \{\mho\} + \later (LX) + \later (LX)$. Then the type of $\run$ becomes
%
\begin{align*} \run : \term{A} \to T(\val{A}) 
    &= \term{A} \to (\bool \to L(\bool \times \val{A})) \\
    &\cong (\term{A} \times \bool) \to L(\bool \times \val{A}).
\end{align*}
%
This is analogous to the usual notion of stateful operational
semantics, in which a term and an initial state are mapped to a final
state and a returned value. Here, $L$ accounts for errors and our two
notions of computational steps.

% Consider the program $M:= \get(M_{t}, M_{f})$ where $M_t :=
% \put_f(\zro)$ and $M_f := \put_t(\suc\, \zro)$. The program reads a boolean from
% the state; if it reads $t$, it writes $f$ and returns $\zro$. Otherwise, it
% writes $t$ and returns $\suc\, \zro$. 
% %
% \[ 
%   \run M = \delta_b(\sem{\get}(
%     \delta_b(\sem{\put_f}(\eta\,\zro)), 
%     \delta_b(\sem{\put_t}(\eta\, (\suc\, \zro))))). 
% \]
% For the state monad above, we have $\sem{\get}\,f\, s = f\, s\, s$, and
% $\sem{\put_s}\,x \,s' = x\, s$, while $\sem{\delta_b}\,x\, s = L.\delta_b(x\,
% s)$. Then simplifying in the above expression, we obtain that $\run M = \lambda
% s.\, L.\delta_b()$

\section{Denotational Semantics}\label{sec:denotational-semantics}

In this section, we discuss how to adapt the denotational model of Giovannini et
al. \cite{gtt-sgdt} to accommodate algebraic effects. After reviewing the key
aspects of their model, we discuss the generalizations necessary to extend their
semantics to accommodate an arbitrary algebraic theory. We additionally correct
a mismatch with step counting between their semantics and the operational
semantics related to the denotation of certain casts.
% \max{limitation is maybe the wrong word. Just say it is a mismatch with the step counting in the operational semantics}.
While this mismatch was not problematic for their result, this change will prove
crucial for us in relating the operational and denotational semantics in Section
\ref{sec:relating-the-semantics}.

% Recall that in their model, pure values were modeled as elements of a
% \emph{predomain}, and effectful computations were modeled as elements of an
% \emph{error domain}, which is a predomain equipped with algebraic operations
% supporting error and computational steps. Our approach will be to generalize the
% notion of error domain to be an algebra of a suitable endofunctor corresponding
% to the signature of effects being modeled. Then the free error domain
% construction will be generalized to construct the free such algebraic structure,
% as an \emph{inductive-guarded interaction tree}.


\subsection{Review of the Prior Denotational Model}

% In this section, we discuss how to build a denotational model of effectful
% gradual typing from scratch in guarded type theory.
%
% Recall that our framework is parameterized by an algebraic theory $\calA$.
%
% Our goal will be to generalize the notion of \emph{error domain} from prior work
% to accommodate these effects.
%
We begin by reviewing the key ingredients in the denotational semantics of
Giovannini et al. \cite{gtt-sgdt}. For further discussion, see their paper. For
the sake of brevity, we will refer to their model as the \emph{prior guarded
model} of gradual typing.

The prior model is based on call-by-push-value \cite{levy99}, a categorical
model that refines Moggi's monadic semantics \cite{MOGGI199155}. In
call-by-push-value, there is a category of value types $\calV$ and pure
morphisms $f : A \to A'$, and a category $\calE$ of computation types $\phi : B
\multimap B'$. A pair of adjoint functors $F \vdash U$ mediates between the two
categories, where $F : \calV \to \calE$ takes $A$ to the type $F A$ of
computations that return values in $A$, and $U : \calE \to \calV$ takes $B$ to
the type $U B$ of suspended computations of type $B$.

In the prior model, the value types are modeled by sets equipped with a partial
ordering relation that models type precision on closed values. However, since
step-insensitive reasoning does not play well with transitivity, each type also
has an additional relation called \textbf{weak bisimilarity}. They call such a
set a \textbf{predomain}. Morphisms of predomains are functions on the
underlying sets that respect both of the relations.
%
The precise definitions are given below:
%
\begin{definition}
A \textbf{predomain} $A$ consists of a set $A$ along with two relations:
\begin{itemize}
    \item A partial order $\ltdyn_A$.
    \item A reflexive, symmetric ``bisimilarity'' relation $\bisim_A$.
\end{itemize}

A \textbf{morphism of predomains} $f : A \to A'$ is a function between the
underlying sets that is both \emph{monotone} (if $x \ltdyn_A x'$, then $f(x)
\ltdyn_{A'} f(x')$), and \emph{bisimilarity-preserving} (if $x \bisim_A x'$,
then $f(x) \bisim_{A'} f(x')$).
\end{definition}

Computation types are equipped with the same two relations, in addition to a
special distinguished error element that is the least element in the ordering,
as well as a distinguished morphism of predomains $\theta_B : \later B \to B$.
%
The precise definitions are given below:
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the
following data:
\begin{itemize}
    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho
    \ltdyn_B x$ for all $x$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
    \item For all $x, y : B$ with $x \bisim_B y$, we have $\theta_B(\nxt\, x)
    \bisim_B y$
\end{itemize}
\end{definition}

A \textbf{morphism of error domains} is a morphism of the underlying predomains
that preserves the error element and the $\theta$ map.
%
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
\circ \nxt$.

In the prior model, type precision is modeled as a kind of heterogeneous
relation on the denoted predomains. More concretely, A \textbf{predomain
relation} $c : A \rel A'$ is a relation between $A$ and $A'$ that is upward
closed in the ordering on $A'$ and downward closed in the ordering on $A$
(no condition is required involving the bisimilarity relations).
%
\begin{definition}
  A \textbf{predomain relation} $c : A \rel A'$ is a relation $c$ between the
  underlying sets of $A$ and $A'$ satisfying:
  \begin{enumerate}
    \item (upward closure): For all $x \in A$ and $y, y' \in A'$, if $x
    \binrel{c} y$ and $y \ltdyn_{A'} y'$, then $x \binrel{c} y'$.
    \item (downward closure): For all $x, x' \in A$ and $y \in A'$, if $x'
    \ltdyn_{A} x$ and $x \binrel{c} y$, then $x' \binrel{c} y$.
  \end{enumerate}
\end{definition}
%
The ``reflexive'' predomain relation is written $r(A)$ and is given by the
ordering $\ltdyn_A$. The relational composition of two predomain relations is
also a predomain relation, and $r(A)$ is the identity for this composition.

Error domain relations are predomain relations that are additionally a
congruence with respect to the operations $\theta$ and $\mho$:
%
\begin{definition}
  An \textbf{error domain relation} $d : B \rel B'$ is a relation $d$ on the
  underlying predomains satisfying
  \begin{enumerate}
     \item ($\mho$ congruence): $\mho_B \binrel{d} \mho_{B'}$.
     \item ($\theta$ congruence): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t (\tilde{x}_t \binrel{d}
     \tilde{y}_t)$, then $(\theta_B (\tilde{x}) \binrel{d} \theta_{B'}
     (\tilde{y}))$. 
  \end{enumerate}
\end{definition}
Note that because $\mho_B$ is a least element this implies the
stronger property $\mho_B \binrel{d} y$ for any term $y: B'$.
%
There is again a reflexive error domain relation $r(B)$ given by the ordering.
The relational composition of two error domain relations in general is not an
error domain relation ($\theta$ congruence in particular is not preserved), but
we instead use an alternative ``free composition'' $dd' : B \rel B''$ that is
inductively generated from the relational composition.

Lastly, term precision is modeled via the notion of \emph{square}, which
captures the logical-relations style relationship between the functions denoted
by two terms. More concretely:
%
\begin{definition}
Let $f : A_i \to A_o$ and $g : A_i' \to A_o'$ be predomain morphisms, and let
$c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$ be predomain relations. There is
a \textbf{predomain square} $f \ltsq{c_i}{c_o} g$ if for all $x \binrel{c_i} y$,
we have $f(x) \binrel{c_o} g(y)$. 

Likewise, an \textbf{error domain square} $\phi \ltsq{d_i}{d_o} \psi$ is defined
similarly, with no extra interaction with the algebraic structure.
\end{definition}
%

\subsection{Generalizing to Algebraic Effects}

In the remainder of this section, we discuss the changes we must make to the
semantics of Giovannini et al. to accommodate arbitrary algebraic effects. The
key question we answer is how to generalize the notion of error domain from
their model to account for the presence of effects besides error and
computational steps.

\subsubsection{Models of Guarded Algebraic Theories in Predomains}

As a first step, we work over an arbitrary guarded algebraic theory that has
suitable operations for error and stepping. Even though our framework is
parameterized by an \emph{ordinary} algebraic theory, everything in this section
applies to the more general setting of guarded algebraic theories. This section
can be seen as analogous to Section \ref{sec:guarded-algebraic-theories}, except
now, instead of defining models of guarded algebraic theories in the category of
sets, we define models in the category of predomains and their morphisms.

Let $\calA = (\Sigma, E)$ be an ordinary algebraic theory (see Section
\ref{sec:guarded-algebraic-theories} for the relevant definitions). Recall that
$\Sigma$ is a signature consisting of a type of operation names $\Op$, where
each $\oper : \Op$ has ordinary arity $\tau_\oper$ and trivial guarded arity.
%
As in the section on operational semantics, let $\calAs = (\Sigma, E) \in
\calA[\mho, \theta]$ be a guarded extension of $\calA$.
%
This means in particular that $\calAs$ has a distinguished error operation $\err
: (0, 0)$ and a distinguished guarded step operation $\theta : (0, 1)$.

In the following, we will need the fact that the induced polynomial endofunctor
$\Sigma : \type \to \type$ extends to an endofunctor on the category of
predomains and their morphisms. In particular, for a predomain $X$, we define
%
\[ \Sigma(X) := \Sigma_{\oper: \Op}, (\tau_\oper \to X) \times (\gamma_\oper \to \later X). \]
%
In the above, the ordering on $\later X$ is defined such that $\tilde{x}
\ltdyn_{\later X} \tilde{y}$ if one time step later, the values are related in
$X$. Likewise for the bisimilarity relation.
%
In addition, the type of functions $\tau_\oper \to X$ inherits a predomain
structure from that of $X$: the ordering and bisimilarity relations are defined
pointwise using the corresponding relations for $X$.

We define the action of $\Sigma$ on morphisms as follows. Given a function $f :
X \to Y$, we define $\Sigma(f) : \Sigma(X) \to \Sigma(Y)$ by
%
\[ \Sigma(f)(\oper, (u, g)) = (\oper, (u', g')), \]
%
where $u' : \tau_\oper \to Y$ is defined by $u' = f \circ u$ and $g' :
\gamma_\oper \to \later Y$ is defined using the functorial action of $\later$ by
$g' = (\later f) \circ g$.

The analogue of error domains in our semantics will then be an algebra $B$ of
the endofunctor $\Sigma$. Using the action of $\Sigma$ on predomains defined
above, an algebra of $\Sigma$ is a predomain $B$ equipped with a morphism of
predomains $\Sigma(B) \to B$. 


% a distinguished error element $\mho$ and a distinguished $\theta_B \colon \later
% B \to B$. 

% As in the definition of an error domain from the denotational model of
% Giovannini et al., we require that $\mho$ is the smallest element in the
% ordering, that $\theta$ respects the relations (i.e., it is a morphism of
% predomains), and that $\delta_B := \theta_B \circ \nxt$ is bisimilar to the
% identity.

% As in the definition of an error domain, we also require that the algebra comes
% with ordering and a bisimilarity relations, that $\mho$ is the smallest element
% in the ordering, that $\theta$ respects the relations (i.e., it is a morphism of
% predomains), and that $\delta_B := \theta_B \circ \nxt$ is bisimilar to the
% identity.


More formally, we make the following definition:
%
\begin{definition}\label{def:sigma-domain} 
  
  Let $\calAs = (\Sigma, E) \in \calA[\mho, \theta]$ as above. An
  \textbf{$\calAs$-domain} is a predomain $B$ with a morphism of predomains
  $\alpha : \Sigma(B) \to B$. In particular, for each $\text{op} \in
  \Sigma.\Op$, we have a morphism of predomains
  %
  \[ \sem{\oper} : (\tau_\oper \to B) \times (\gamma_\oper \to \later B) \to B. \]
  %
  
  Furthermore, each equation $e = (\Gamma_e, l_e, r_e)$ in $E$ holds when
  interpreted in $B$.

  Lastly, the predomain $B$ has a distinguished element $\mho_B$ such that
  $\mho_B \ltdyn_B x$ for all $x \in B$, along with a distinguished morphism of
  predomains $\theta_B : \later B \to B$, such that if $x \bisim_{B} y$, then
  $\delta_B \bisim_{B} y$ where $\delta_B := \theta_B \circ \nxt$.
\end{definition}
%
As with ordinary error domains, there are corresponding notions of morphisms,
relations, and squares of $\calAs$-domains. Before introducing these, we first
discuss the action of the endofunctor $\Sigma$ on predomain morphisms and
predomain relations.

Let $c : A \rel A'$ be a predomain relation. The action of the endofunctor
$\Sigma$ on $c$, written $\Sigma(c)$, is a predomain relation between
$\Sigma(A)$ and $\Sigma(A')$, defined as follows:
%
First, note that any predomain relation $c : A \rel A'$ extends to a predomain
relation $\tau \to c$ between $\tau \to A$ and $\tau \to A'$ defined pointwise,
i.e., $f \binrel{(\tau \to c)} f'$ iff for all $s : \tau$ we have $(f s)
\binrel{c} (f' s)$.
%
Similarly, given $c : A \rel A'$ we define $\later c$ such that $\tilde{x}
\binrel{(\later c)} \tilde{y}$ iff after one step passes, the left and right
sides (now in $A$ and $A'$ respectively) are related in $c$.
%
We now define $(\oper, (u, g)) \binrel{(\Sigma(c))} (\oper', (u', g'))$ iff
$\oper = \oper'$ and $u \binrel{(\tau_\oper \to c)} u'$ and 
$g \binrel{(\gamma_\oper \to \later c)} g'$.


Now we can introduce the definitions of morphisms and relations of
$\calAs$-domains.
%
\begin{definition}\label{def:sigma-domain-mor-rel-sq} Let $B$ and $B'$ be
  $\calAs$-domains. A \textbf{morphism of $\calAs$-domains} from $B$ to $B$' is a
  morphism $\phi$ of the underlying predomains that preserves the algebraic
  structure.
  %
  Spelling this out concretely, we have:
  %
  $\phi (\sem{\oper}_B(u, g)) = \sem{\oper}_{B'}((\phi \circ u), ((\later \phi) \circ g))$,

  A \textbf{relation of $\calAs$-domains} $d : B \rel B'$ is a relation $|d|$ of
  the underlying predomains that is a congruence with respect to the algebraic
  structure. In particular:
  %
  If $x \binrel{(\Sigma(|d|))} y$, then $\sem{\oper}_B(x) \binrel{|d|} \sem{\oper}_{B'}(y)$.

  A \textbf{square of $\calAs$-domains} $\phi \ltsq{d}{d'} \phi'$ is defined as a
  square of the underlying predomains. 
\end{definition}

\begin{remark}
  Spelling out the definition of morphism for the distinguished guarded stepping
  operation $\theta : \later B \to B$, we get that
  %
  $\phi(\theta_{B}(\tilde{x})) = \theta_{B'}((\later \phi)(\tilde{x}))$.

  The requirement that $\mho_B$ be related to $\mho_{B'}$ in the definition of
  relation may seem weaker than the corresponding requirement in the definition
  of a relation of error domains in the previous section, where we required that
  $\mho_B \binrel{|d|} y$ for \emph{all} $y \in B'$. However, the former implies
  the latter using the fact that predomain relations are closed under the
  ordering on both sides and that $\mho_{B'} \ltdyn y$ for all $y$.
\end{remark}

The usual relational composition of $\calAs$-domain relations is not in general
a $\calAs$-domain relation. This is because congruence with respect to guarded
operations (such as $\theta : \later B \to B$) is not preserved by the
relational composition. Thus, we instead define a free composition $dd' : B \rel
B''$ that is inductively generated from the relational composition. See the
paper of Giovannini et al. for details \cite{gtt-sgdt}.
% See the Appendix for the precise definition.
% TODO cite appendix.

\subsubsection{Lifting the Free Models to Predomains}\label{sec:extending-to-predomains}

Let $\calA$ be an algebraic theory, and let $\calAs \in \calA[\mho, \theta^s]$
be a guarded extension with error and step operations. Suppose the free models
of $\calAs$ are given by a functor $F$ defining a strong adjunction between sets
and $\calAs$-models. In analogy with the prior guarded denotational model of
graduality, we seek to lift this adjunction to a strong adjunction between
predomains and $\calAs$-domains. To do so, we need to define the error ordering
and weak bisimilarity relation on $T(X)$. Because we require the error ordering
to be antisymmetric, we cannot define it generically --- we cannot guarantee
that the resulting relation will be antisymmetric. Instead, the construction of
the denotational semantics takes this relational data as input. Concretely, the
data consists of the definition of the weak bisimilarity relation, and the
lifting of a predomain relation $c : A \rel A'$ to a $\calAs$-domain relation
$T(c)$. The latter is required to have the property that given a reflexive,
antisymmetric relation $\ltdyn_A$, the lifting is also reflexive and
antisymmetric. In addition, it must satisfy the following universal property: if
we have a predomain square $f \ltsq{c}{Ud} g$ where $f : A \to UB$ and $g : A'
\to UB'$ and $d : B \rel B'$ is a $\calAs$-domain relation, then there is a
square $f^* \ltsq{T(c)}{d} g^*$ where $f^*$ denotes the extension of $f$ to a
homomorphism $T(A) \to B$.
%
Lastly, we require that the algebraic operations on $T(A)$ respect the ordering
and bisimilarity relations.

\subsection{Perturbations and Quasi-Representable Relations}

We next review how the model of Giovannini et al. validates the graduality
property. In the New-Licata denotational model of graduality on which Giovannini
et al. based their guarded model, the behavior of casts was encoded in the
semantics by \emph{representable relations}. In prior work, these were
sufficient for interpreting the four cast rules crucial for graduality. At a
high level, type precision $c : A \ltdyn A'$ is modeled by representable
relations, i.e., relations that behave like the graph of a function, e.g., $x
\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A'} y$
\cite{shulman2007}.

However, as Giovannini et al. showed, the notion of representability cannot be
applied unchanged to the guarded setting because of a fundamental tension
between transitivity and step-insensitive reasoning \cite{gtt-sgdt} (see Section
4 of their paper for details). As a result, the authors worked with a transitive
but step-sensitive ordering where terms must be in lock-step. They weakened the
representability requirement to a notion that they called
\emph{quasi-representability}, which is essentially representability up to an
explicit delay. They then developed a theory of these delays, which they called
\emph{syntactic perturbations}.

More concretely, in the prior guarded semantics, value types denote predomains
along with a monoid of syntactic perturbations $M_{A}$ and an interpretation
homomorphism $i_A : M_{A} \to \semptb{A}$, where $\semptb{A}$ is the set of
endomorphisms on $A$ that are weakly bisimilar to the identity. Computation
types were defined analogously. The idea is that the perturbations mimic the
steps taken by casts, allowing the perturbed term to be in lock-step with a
cast.
% and because they are specified as syntactic data, it is possible to associate 

Then to model type precision, the authors required that each predomain relation
$c : A \rel A'$ is quasi-left-representable, and $Fc$ is
quasi-right-representable. Likewise, error domain relations $d : B \rel B'$ are
required to be quasi-right-representable and $Ud : UB \rel UB'$ must be
quasi-left-representable.
%
Lastly, in order to make the notions of quasi-representability compositional,
the authors additionally require that the relations must interact with the
perturbations on either side, in that given $c : A \rel A'$, it is possible to
push a perturbation on $A$ to one on $A'$ and pull a perturbation on $A'$ to one
on $A$.
%
On the other hand, morphisms are not required to interact with the
perturbations.
% TODO more detail on the definition of quasi-representability?

% It turns out\max{Don't start a paragrpah with ``It turns out''. Say something more active like ``their proof cleanly generalizes''}
The construction of perturbations and quasi-representable relations is
independent of the details of the adjunction between predomains and (our
generalized notion of) error domains. This means that the constructions used in
the model of Giovannini et al. will generalize cleanly to our setting without
any major modifications.
%
Details as to the definitions of the perturbations associated with each of the
type constructors can be found in the Appendix. % TODO cite appendix
%
The intuitive reason for why no modifications are needed is that the casts do
not cause side effects other than stepping or error. Thus, the perturbations for
$F(A)$ can still be defined as $\mathbb{N} \oplus M_A$.

\subsection{Adjusting the Term Semantics}\label{sec:term-relational-semantics}

% ...we need to make an important technical modification to the model of Giovannini
% et al.
In addition to generalizing error domains, we need to make an important
technical modification to our model concerning the behavior of casts. This change is
motivated by our goal of relating the denotational and operational semantics via
an adequacy result. In the prior model, the denotations of certain casts
involves additional perturbations (i.e., computational steps) in order to place
terms involving those casts in lock-step. These additional steps ultimately did
not matter for the final result, since the graduality property is
step-insensitive. However, intensionally, these casts behave differently from
the corresponding casts without the extra steps.

This difference in intensional behavior presents a difficulty when relating
their denotational model to an operational semantics: in order for the two
semantics to be in lock-step, we must define the operational semantics in such a
way that the casts take additional steps just as they do in the denotational
semantics. But this requirement is not desirable, as it restricts the kinds of
operational semantics to which our framework applies to those that happen to
feature this non-standard cast semantics. Users of our framework would need to
contort their operational semantics into this form if they would like to apply
our adequacy result.

Our remedy for this situation is to note that there is an additional degree of
freedom in the model of Giovannini et al. that up to this point has remained
unexploited. Leveraging this, we define two \emph{different} cast semantics, one
where the casts involve perturbations, and one where they do not. The former is
used in constructing the semantic proofs of the cast axioms, where being in
lock-step is crucial. The latter is what we use for the denotational term
semantics of casts, allowing the denotations of casts to be in lock-step with
the casts in a typical operational model, i.e., where the casts do not take
additional steps. While the intensional behavior of these two kinds of casts are
distinct, they are extensionally related in that they are \emph{weakly
bisimilar}, which is sufficient for the final proof of graduality.

For concreteness, term precision $M \ltdyn N : A$ is modeled in the prior
semantics by $\sem{M} \ltbisim \sem{N}$ where
%
\[ f \ltbisim g := \exists f', g'. f \bisim f' \ltdyn g' \bisim g. \]
%
That is, we do not require that $f$ and $g$ are in lock-step; we are allowed to
use any $f'$ bisimilar to $f$ and $g'$ bisimilar to $g$ such that $f'$ is
lock-step related to $g'$ in the error ordering.
%
This is useful because the downcast associated to the composition of type
precision $c : A \ltdyn A'$ and $c' : A' \ltdyn A''$ is a morphism from
$F(\sem{A''})$ to $F(\sem{A})$, but in the prior model it is not simply the
composition of the downcast morphisms for $c$ and $c'$. Instead, an additional
perturbation is inserted in order to make the relational squares line up in the
proof of quasi-representability. More details can be found in the Appendix of
their paper \cite{gtt-sgdt-extended-version}.

On the other hand, observe that we could instead define the downcast morphism
for the composition $cc'$ to be the composition of the corresponding downcasts,
without the additional perturbation, and then note that the resulting morphism
is \emph{weakly bisimilar} to the version with the additional perturbation. This
is sufficient for graduality, since the additional perturbation will be absorbed
by the weak bisimilarity in the definition of $f \ltbisim g$.

% Using the above definition for term precision, we see that the additional
% perturbations in the cast corresponding to a composition will be absorbed by the
% weak bisimilarity.

With this observation in mind, we augment the denotation of type precision in
the prior denotational model by associating to each value relation $c$ in the
semantics two morphisms: a ``relational'' embedding morphism $e$ such that $c$
is quasi-left-representable by $e$, and a ``term'' embedding morphism $e' \bisim
e$. We similarly associate two projection morphisms to each computation relation
$d$. We use the term versions of the embedding and projection morphisms as our
denotations of upcasts and downcasts, and we use the relational versions
internally in the proof of graduality.
%
The result is that our term semantics for casts no longer involves unnecessary
perturbations that were inserted to make the graduality proof work.

% TODO cite appendix
Details on the changes made to the model can be found in the Appendix.
\max{this section is a good candidate for being squashed for space since a lot of it is in the previous paper/unchanged}

\section{Relating the Operational and Denotational Semantics}\label{sec:relating-the-semantics}

Having specified the operational and denotational semantics, we now seek to
prove a theorem relating them. Our goal is to show that the denotational model
is \emph{adequate} for the operational semantics, that is, if the denotations of
two terms are equal, then their operational behavior is the same.

More concretely, in Section \ref{sec:operational-semantics}, we defined a
function $\run : \term{A} \to T_{\mho sb} (\val{A})$ taking a closed term to its
final behavior. Then in Section \ref{sec:denotational-semantics} we defined the
denotation of a term as a function $\sem{\cdot} : \term{A} \to T_{\mho
s}(\sem{A})$. Our goal is to relate the operational and denotational semantics.
In particular, we seek to show that for closed terms of type $\nat$, we have
%
\[ \run(M) \equiv_b \sem{M}, \]
%
where $\equiv_b$ is a relation between $T_{\mho sb}(\mathbb{N})$ and $T_{\mho
s}(\mathbb{N})$ that ignores the bureaucratic delays in its first argument.

To that end, we will define a logical relation between the denotations of
(closed) terms of type $A$ and syntactic terms of type $A$:
%
\[ \calE_A : \Rel {T_{\mho s}\sem{A}} {(\term A)}, \]
%
and then $\equiv_b$ is defined as $\calE_\nat$ since we have $\sem{\nat} = \mathbb{N}$.

\subsection{The Logical Relation}

% Note: here we *do* care about the fact that the theories are built using
% sum/tensor, not merely that they have error and stepping operations.

Before defining the logical relation, we will need the notion of a
\emph{heterogeneous lifting} which takes a relation $R : X \rel Y$ and lifts it
to $\bbt(R) : T_{\mho s}(X) \rel T_{\mho sb}(Y)$. The lifted relation relates
bureaucratic steps in its right argument to the identity in its left argument.
We define the lifted relation as the least relation that (1) extends $R$, (2) is
a congruence with respect to the operations in $\calA_{\mho s}$ (i.e., the
operations of $\calA$, error, and \emph{true} stepping), and (3) relates the
identity in $T_{\mho s}(X)$ to $\delta_b$ in $T_{\mho sb}(Y)$. The precise
definition is given in Figure \ref{fig:relational-lifting}.

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=RetRet]
    { x \binrel{R} y }
    { \eta^{T_{\mho s}}(x) \binrel{(\bbt(R))} \eta^{T_{\mho sb}}(y) }

    \inferrule*[right=OpOp]
    { \oper \in \calA_{\mho s}.\Op 
      \and \forall s.\, (u\, s) \binrel{(\bbt{R})} (u'\, s) \\
      \and \forall s.\, (g\, s) \binrel{(\later(\bbt{R}))} (g'\, s)  
    }
    { (T_{\mho s}(X).\oper(u, g)) \binrel{(\bbt{R})} (T_{\mho sb}(Y).\oper(u', g')) }

    \inferrule*[right=IdStep]
    { \bar{x} \binrel{(\bbt(R))} \bar{y} }
    { \bar{x} \binrel{(\bbt(R))} \delta_b(\bar{y}) }

  \end{mathpar}

  \caption{Relational Lifting that Ignores Bureaucratic Steps}
  \label{fig:relational-lifting}
\end{figure}


The logical relation is defined in Figure \ref{fig:logical-relation}. We define
a \textbf{value relation}
%
$\calV_A : \Rel{\sem{A}} {\val{A}}$,
%
and an \textbf{expression relation} 
%
$\calE_A : \Rel{T_{\mho s}\,\sem{A}}{\term{A}}$.
%
While logical relations are typically defined by induction on the type $A$, the
case for $\dyn$ is generated by a combination of induction and guarded
recursion. Note that while the product case for $\dyn$ refers to the value
relation at the larger type $\dyn \times \dyn$, this is merely syntactic sugar;
to be completely rigorous, we would simply \emph{define} the relation
$\calV_{\dyn \times \dyn}$ in an analogous manner to the definition for the
product case $\calV_{A \times A'}$.
% TODO explain this further?


\begin{figure}
  \begin{mathpar}
    \begin{array}{rcl}
      \calV_\nat (n, V_n) &:=& (n = V_n) \\
      \calV_{A_i \ra A_o} (f, V_f) &:=& 
        \forall (x \binrel{\calV_{A_i}} V).\, (f\, x) \binrel{\calE_{A_o}} (V_f\, V) \\
      \calV_{A \times A'} ((x, y), V_\times) &:=& \exists V_1, V_2.
        (V_\times = (V_1, V_2)) 
        \wedge (x \binrel{\calV_{A}} V_1) 
        \wedge (y \binrel{\calV_{A'}} V_2) \\
    \end{array}

    \inferrule*[right=DynNat]
    { n = V_n }
    { \vrel{dyn}{\inat(n)}{\tagnat {V_n}} }

    \inferrule*[right=DynProd]
    { \vrel{\dyn \times \dyn}{(d_1, d_2)}{V_{d\times}} }
    { \vrel{\dyn}{\itimes(d_1, d_2)}{ \tagtimes{V_{d\times}}} }

    \inferrule*[right=DynFun]
    { \later(\vrel{\dyn \ra \dyn} {f} {V_f}) }
    { \vrel{\dyn}{\iarr(f)}{\tagarr{V_f}} }


  \begin{array}{rcl}
      \erel {A} {\bar{x}} {N} &:=& \liftvrel{A}{\bar{x}}{(\run N)} % \bar{x} \binrel{(\bbt(\calV_{A}))} (\run{N})
  \end{array}

  \end{mathpar}

  \caption{Logical Relation}
  \label{fig:logical-relation}
\end{figure}

The logical relation satisfies the following properties.
%
First, we have \emph{anti-reduction}:
%
\begin{lemma}[Anti-Reduction]\label{lem:anti-reduction}

  Let $\cdot \vdash M : A$, and suppose $M \stepsin{k} \cpure{M'}$. If
  $\erel{A}{\xbar}{M'}$, then $\erel{A}{\xbar}{M}$.

\end{lemma}

Next, we have a semantic bind lemma for the logical relation, which allows us to
replace arbitrary terms in an evaluation context with values.
%
\begin{lemma}[Monadic Bind]\label{lem:monadic-bind}
  
  % Suppose for all values $V, W$ that $\vrel{A}{V}{W}$ implies
  % $\erel{B}{\phi(V)}{(E[W])}$. Then, for all $\erel{A}{\bar{x}}{N}$, we have
  % $\erel{B}{(\phi(\bar{x}))}{(E[N])}$.
  Let $\phi : T_{\mho s}(\sem{A}) \to T_{\mho s}(\sem{B})$, and let $\bullet : A \vdash E : B$.
  %
  Suppose for all values $x, V$ that $\vrel{A}{x}{V}$ implies
  $\erel{B}{\phi(T_{\mho s}.\eta(x))}{(E[V])}$.
  %
  Then for all $\bar{x} \in T_{\mho s}(\sem{A})$ and terms $N$ of type $A$ such that
  $\erel{A}{\bar{x}}{N}$, we have $\erel{B}{(\phi(\bar{x}))}{(E[N])}$

\end{lemma}


\subsection{Fundamental Lemma}

% TODO define \grel{\Gamma}

% $\gambar$ maps $x : A \in \Gamma$ to $\sem{A}$ and $\gamma'$ maps $x : A \in \Gamma$ to $\val{A}$

We begin by defining the extension of the value relation to contexts.
Concretely, we define $\grel{\Gamma}{\gambar}{\gamma'}$, where $\gambar :
\sem{\Gamma}$ and $\gamma'$ is a mapping of variables $x : A \in \Gamma$ to
$\val{A}$, as the pointwise extension of $\vrel{}{}{}$ to the elements in
$\gambar$ and $\gamma'$.
%
% The definition is by induction, with rules
% %
% $\grel{\bullet}{\_}{\_} := \top$, and
% %
% $\grel{\Gamma, x : A}{\bar{x} :: \bar{xs}}{[x \mapsto V, xs]} :=
%   \vrel{A}{\gambar(x)}{\gamma'(x)} \wedge \grel{\Gamma}{\bar{xs}}{xs}$.
%
Now we define the extension of the logical relation to morphisms $\fbar :
\sem{\Gamma} \to T_{\mho s}(\sem{A})$ and open terms $\Gamma \vdash M : A$, as
follows:
\[ 
  \Gamma \vDash \fbar \sim M : A := 
    \forall (\grel{\Gamma} {\gambar} {\gamma'}).\,
      \erel{A} {(\fbar\, \gambar)} {M[\gamma']}.
\]
%
We now state the fundamental lemma for the logical relation, which implies the
result that $\run(M) \equiv_b \sem{M}$:
%
\begin{lemma}[Fundamental Lemma]
  Let $\Gamma \vdash M : A$. Then $\Gamma \vDash \sem{M} \sim M : A$.
\end{lemma}
%
For the proof of this lemma, see the Appendix. % TODO cite appendix
%
The above implies the adequacy of the denotational semantics with respect to the
operational semantics:
\begin{theorem}[Adequacy]
  Let $\cdot \vdash M : \nat$. Then we have $\run{M} \equiv_b \sem{M}$.
\end{theorem}

% \subsection{Summary of Results So Far}

% Logical relation
% Fundamental Lemma
% Adequacy of the term model
% Adequacy of the relational model


% To this end, we will define a logical relation between (closed) syntactic terms
% of type $A$ and their denotations as elements of $F\sem{A}$, where $F$ is the
% functor constructing the free error-effect algebra as discussed in Section
% \ref{sec:denotational-semantics}.


\section{Globalization and Extensional Collapse}\label{sec:globalization}

\begin{comment}
If the provided algebraic theory has no equations, then the free-model monad for
the theory $T(X)$ is isomorphic to the type of unquotiented terms with variables
in $X$.

To motivate the forthcoming definitions, we begin with a simple example where
the only effects are error and computational steps. Consider the guarded
algebraic theory $\calA = \error + \steptheory$ (see Section
\ref{sec:guarded-algebraic-theories} for the definitions). In particular, note
that there are two operations $\mho : (0, 0)$ and $\theta : (0, 1)$. The
free-model monad for this theory is the unique type $T(X)$ solving the
guarded-recursive equation
%
\[ T(X) \cong X + 1 + \later (T(X)). \]

We now review the notion of \emph{clocks}, which are a tool allowing us to
extract coinductive definitions from guarded ones. Clocks were introduced by
Atkey and McBride \cite{atkey-mcbride2013}. The idea is that all guarded
constructions are indexed by a clock variable $k$, e.g., $\later^k A$ is the
type of $A$-values available after one time step on clock $k$.
%
The key aspect of clocks is that universal quantification over clocks gives a
means to turn a guarded definition into a coinductive one
\cite{atkey-mcbride2013, kristensen-mogelberg-vezzosi2022}.


Define $T^{gl}(X) := \forall k. T^k(X)$. Then provided $X$ satisfies a technical
condition known as \emph{clock-irrelevance}, we have
%
\begin{align*}
  T^{gl}(X) &= \forall k. (X + 1 + \later T^k(X)) \\
  &\cong \forall k. X + \forall k. 1 + \forall k. \later T^k(X) \\
  &\cong X + 1 + \forall k. T^k(X).
\end{align*}
%
Moreover, $T^{gl}(X)$ is the final coalgebra for the functor $F_X(Y) = (X + 1) +
Y$. This follows from a theorem of Kristensen et al.
\cite{kristensen-mogelberg-vezzosi2022}. The final coalgebra for $F_X$ can be
expressed using Capretta's \emph{delay monad} $D(X)$, which is defined as the
final coalgebra of the functor $F_X(Y) = X + Y$ \cite{lmcs:2265}. Elements of
$D(X)$ can be thought of as computations that may either return a value in $X$,
or take an observable computational step. Such computations may run forever,
which makes $D(X)$ a constructive method for modelling non-termination in type
theory.

Altenkirch et al. \cite{Altenkirch-Danielsson-Kraus-2017} show that the delay
monad $D(X)$ is equivalent to the type of \emph{monotone sequences} $\mathbb{N}
\to (X + 1)$ (see Section 4.1 of their paper). By a monotone sequence, we mean a
function $f : \mathbb{N} \to (X + 1)$ satisfying that if $f(n) = \inl x$, then
$f(n+1) = \inl x$. The intuition is that while we cannot decide in general
whether an arbitrary computation terminates or runs forever, for any
\emph{finite} amount of ``fuel'' $n$, we can ask whether it has terminated in
$n$ steps or fewer. If it has, then $f$ tells us the value with which the
computation terminated; if not, we are simply told that it is still running.
Monotonicity says that these answers must be consistent: if the computation
terminates in $n$ or fewer steps with result $x$, then it certainly also
terminates in $n+1$ or fewer steps with the same result $x$.

Observe that the delay monad is an intensional representation of
potentially-nonterminating computations. That is, two computations that return
the same value but take a different number of steps are considered distinct.
Since the statement of graduality is oblivious to the number of steps a program
takes, the delay monad is not suitable as a denotational domain for closed
programs. To remedy this situation, one may consider the delay monad quotiented
by weak bisimilarity. However, in order to show that the resulting type is a
monad in type theory, it is believed that the axiom of countable is required
\cite{TODO}.


To generalize the situation to arbitrary effects, we assume that our theory
$\calA$ contains no equations. Then we consider the theory $\calA + \th{\mho} +
\th{\step}$. Then as above, we form the free-model monad for this theory,
and consider the globalized version by performing clock quantification.
%
The analogue of the monotone sequences for the simple setting of error and
stepping will be functions that take a generalized notion of fuel and return a
finite trace of the computation.
\max{need to update this section to focus on the traces/trees not lift and certainly not clocks}
\end{comment}

The last step of the graduality proof involves using the results obtained in the
prior sections to derive extensional properties of the semantics that do not
mention any features of guarded type theory. The result is a statement about the
operational behavior of terms that can be made in ordinary dependent type
theory. Furthermore, the statement is step-insensitive, in that it is invariant
under finite differences in number of steps taken by a program.
%
The process we will follow here involves globalization \cite{atkey-mcbride2013},
which is a process that takes a construction in guarded type theory and turns it
into a coinductive object in ordinary type theory. Not all constructions in
guarded type theory can be globalized into coinductive definitions, but the
specific definitions we will be globalizing do satisfy the required technical
condition.
% TODO: mention the notion of a functor commuting with clock quantification.

At a high level, the goal of this last step of the graduality proof is to define
a function transforming elements of the (guarded) free-model monad at base type
$T(\mathbb{N})$ into elements of a type of extensional effectful computations
$E$ that (provided they terminate) return natural numbers. Then we show that
this process maps terms related in the guarded extensional error ordering of
$T(\mathbb{N})$ to extensional computations related in $\ltbisim_E$, where the
latter is a partial order on $E$.

Our approach is as follows. We first perform globalization in guarded type
theory, on both the free-model monad $T(\mathbb{N})$ as well as its associated
error ordering and weak bisimilarity relations. Though the results of
glboalization are coinductive and do not utilize guarded type theory, they are
still \emph{intensional}, in that computations that are weakly bisimilar (i.e.,
differ by a finite number of steps) are considered distinct elements of the
globalized free-model monad. Thus, our final step is to perform a ``collapse'',
the result of which describes the \emph{extensional} behavior of a term. More
concretelty, the extensional collapse is a function $\col : T^{gl}(\mathbb{N})
\to E$, where $T^{gl}$ is the globalization of the free-model monad $T$. We then
require that the global versions of strong error ordering and weak bisimilarity
imply the relation $\ltbisim_E$.
%
More details on this stage of the proof and how it fits into the overall
structure of the graduality proof are given in Section
\ref{sec:graduality-shemata}. 

We are not aware of any general theorem to globalize free algebras for
an arbitrary guarded algebraic theory $\calA$. However, we have
identified two constructions that allow us to construct many examples:
free theories (those with no equations) provide a base case, and then
well-behaved theory/monad transformers allow us to construct more
complex examples that do contain equations.
%
These results allow us to build up the free-model monad for a
suitably-constructed algebraic theory in a compositional manner. In particular,
if the innermost monad is the free model-monad for a finitary free theory, then
we can globalize the monad and apply the sequence of monad transformers to the
result.
%
While we do not support the usual theories of non-determinism or probabilistic
computation since they involve equations and do not arise from one of our theory
transformers, we \emph{do} support their free variants. This is a reasonable
alternative, as the graduality result for the free theories --- a result about
the behavior of programs --- is not invalidated by considering equations.

% \max{This was pretty weaselish before I tried to make it punchier. Please review Eric.}

% \max{I feel like free theory should come first because it's the base
%   case.}

% \max{We should also comment on theories we don't support and that
%   taking the free theory without equations is a somewhat reasonable
%   alternative in those cases (non-det/probability)}.


\begin{comment}
One example of this phenomenon involves the theory of exceptions:
%
\begin{lemma}[Error Monad Transformer]\label{lem:error-monad-transformer} Let
$\calA$ be a guarded algebraic theory. If $T_\calA$ is the free-model monad for
$\calA$, then the free-model monad for the theory $\calA + \th{\mho}$ is given
by
%
\[ T'(A) := T(A + 1). \] 
\end{lemma}

Another example of this phenomenon involves the theory of global mutable state:
%
\begin{lemma}[State Monad Transformer]\label{lem:state-monad-transformer}
Let $\calA$ be a guarded algebraic theory. Suppose that $\calA$ has an operation
$\theta$ of arity $(0, 1)$. If $T_\calA$ is the free-model monad for $\calA$,
then the free-model monad for the theory $\calA \otimes \th{\text{State}_S}$ is
given by
%
\[ T'(A) := S \to T_\calA(S \times A). \]
%
\end{lemma}
\end{comment}

% These results allow us to build up the free-model monad for a
% suitably-constructed theory in a compositional manner. In particular, if the
% innermost monad is the free model-monad for a finitary free theory, then we can
% globalize the monad and apply the sequence of monad transformers to the result.

% Such results allow us to build up the free-model monad in a compositional
% manner. If the innermost monad is the usual guarded lift monad $L(X) \cong X +
% \later L(X)$, then we can then globalize the overall monad and use known results
% about the globalization of the guarded lift monad to obtain a characterization
% of the combined monad.


\section{Summary and Instantiations of the Framework}\label{sec:framework}
In this section, we bring together the various steps of the framework and
instantiate it with some examples corresponding to familiar algebraic theories.

\subsection{Graduality Schemata}\label{sec:graduality-shemata}

We bundle the required data for all the steps of our construction into a
definition we call a graduality schema, defined below:
%
\begin{definition}[$\calA$-Graduality Schema]
Let $\calA$ be an ordinary, finitary algebraic theory. An
\textbf{$\calA$-graduality schema} consists of the following data:
%
\begin{enumerate}[label=\textbf{GS.\arabic*}, ref=GS.\arabic*]

\item \label{itm:gs-guarded-extension} Guarded extensions $\calAs \in
\calA[\mho, \theta^s]$ and $\calAsb \in \calAs[\theta^b]$.

\item \label{itm:gs-free-models} A chosen presentation of the free models for
$\calAs$, given by an adjunction $F \dashv U$ between sets and models of
$\calAs$. Let $T := UF$ denote the associated free-model monad on Set.
% Likewise for $\calAsb$. Let $\Ts$ and $\Tsb$ be the respective monads.

\item \label{itm:gs-lifting} A lifting of the adjunction for $\calAs$ to a
strong adjunction between the double categories of predomains and
$\calAs$-domains (see Section \ref{sec:extending-to-predomains} for details).
Let $\That := UF$ be the associated monad on predomains. Being a lifting means
in particular that for any predomain $A$, we have $|\That(A)| = T(|A|)$ with $T$
as above.

%and let $\That^{gl}(A) := \forall k.\That^k(A)$.

\item \label{itm:gs-globalization} For globalization: $T$ commutes with clock
quantification, i.e., the canonical map $T(\forall k.X[k]) \to \forall k.
T(X[k])$ is an isomorphism.

\item \label{itm:gs-collapse} For the extensional collapse: A family of types
$E(A)$ along with functions $\col : T^{gl}(A) \to E(A)$.

\item \label{itm:gs-ext-ord} A partial order $\ltbisim$ on each $E(A)$, such
that $x \ltdyn^{gl} y$ and $x \bisim^{gl} y$ imply $\col{x} \ltbisim \col{y}$.
\end{enumerate}
\end{definition}

% In item \ref{itm:gs-guarded-extension}, the guarded extension

Given this data, we construct the operational and denotational semantics, and
prove operational graduality. We proceed as follows:
%
\begin{enumerate}
  \item We define the small-step operational semantics as in Section
  \ref{sec:small-step-semantics}, and then define the $\run$ function as
  discussed in Section \ref{sec:big-step-semantics}. This uses the guarded
  extensions $\calAs$ and $\calAsb$ and their associated free-model monads $\Ts$
  and $\Tsb$.

  \item We define a denotational semantics as described in Section
  \ref{sec:denotational-semantics}, using the provided adjunction between
  predomains and $\calAs$-domains (item \ref{itm:gs-lifting}). The perturbations
  and quasi-representability are the same as in the prior guarded model. The
  definition of the dynamic type is also as before, $D \cong \mathbb{N} + D
  \times D + \later U (D \to F D)$.

  \item We establish adequacy of the denotational semantics with respect to the
  operational semantics as described in Section \ref{sec:relating-the-semantics}:
  %
  \[ \run M \equiv_b \sem{M}. \]
  %
  \item Now, to prove graduality for the operational semantics, we start by
  assuming $M \ltdyn N : \nat$. Then by adequacy and guarded denotational
  graduality, we have
  %
  \[ \exists f,g.\, \run M \equiv_b \sem{M} \bisim f \ltdyn g \bisim \sem{N} \equiv_b \run N. \] 
  %
  \item We map the bureaucratic delays in $\run M$ and $\run N$ to true delays. 
  Then $\equiv_b$ turns into $\bisim$:
  %
  \[ d(\run M) \bisim \sem{M} \bisim f \ltdyn g \bisim \sem{N} \bisim d(\run N). \]   
  %
  \item We perform globalization and extensional collapse. First, we globalize:
  \[ \gl(d(\run M)) 
    \bisim^{gl} \gl(\sem{M}) 
    \bisim^{gl} \gl(f) 
    \ltdyn^{gl} \gl(g) 
    \bisim^{gl} \gl(\sem{N}) 
    \bisim^{gl} \gl(d(\run N)). \]
  %
  Then taking the extensional collapse, we obtain
  %
  \[ \col(\gl(d(\run M))) \ltbisim \col(\gl(d(\run N))), \]
  %
  where $\col$ is as described in item \ref{itm:gs-collapse} and $\ltbisim$ is
  as described in item \ref{itm:gs-ext-ord}.
\end{enumerate}


\section{Graduality for Free Theories}

\max{TODO: find the right place to put this}

Consider the case that our algebraic theory is a free theory generated
by some signature, i.e., a set $\Sigma$ of operation names with
arities $\ar : \Sigma \to \mathbb N$. Then we can form the type of
finite trees with node types from the signature:

\[ \textrm{FinTree}_\Sigma X = \mu T. X + ((\sigma : \Sigma) \times T^{\ar(\sigma)}) \]

Then the globalization of a free algebra for this theory extended with
$\mho, \theta$ is a certain kind of coinductive ``interaction tree'':

\[ \textrm{ITree}_\Sigma X = \nu D. \textrm{FinTree}_\Sigma(X + \{ \mho \} + D) \]

For clarity, the $X$ case of a $\textrm{FinTree}X$ will be called
$\textrm{leaf}(x)$ and the $X$ case of $X + \{ \mho \} + D$ in
$\textrm{ITree}_\Sigma X$ will be called $\eta(x)$.

The reason for the interleaving of $\textrm{FinTree}_\Sigma$ is that
because only the $\theta$ operation is guarded, there can be only
finitely depth of operations from the signature unless they are
interrupted by a $\theta$.
We define $M \leadsto N$ to mean that $M = \textrm{leaf}(\delta(N))$,
representing that $M$ ``steps'' to $N$ in one internal ($\tau$) step.

Then the globalization of the strict error ordering $\sqsubseteq$ and
weak bisimilarity $\bisim$ for ITrees imply the following
coinductively generated relations on $\textrm{ITree}_\Sigma X$:

\begin{mathpar}
  \inferrule{}{\eta x \sqsubseteq \eta x}

  \inferrule{}{\err \sqsubseteq N}

  \inferrule{M \sqsubseteq N}{\delta(M) \sqsubseteq \delta(N)}

  \inferrule{\forall r \in \ar(\sigma). f(r) \sqsubseteq g(r)}{\sigma(f) \sqsubseteq \sigma(g)}
\end{mathpar}

\begin{mathpar}
  \inferrule{N \leadstomany \eta x}{\eta x \bisim N}\and
  \inferrule{M \leadstomany \eta x}{M \bisim \eta x}\and
  \inferrule{N \leadstomany \err}{\err \bisim N}\and
  \inferrule{M \leadstomany \err}{M \bisim \err}\and
  \inferrule{N \leadstomany \sigma(g)\and\forall r \in \ar(\sigma). f(r) \bisim g(r)}{\sigma(f) \bisim N}\and
  \inferrule{M \leadstomany \sigma(f)\and\forall r \in \ar(\sigma). f(r) \bisim g(r)}{M \bisim \sigma(g)}\and
  \inferrule{M \bisim N}{\delta(M) \bisim \delta(N)}
\end{mathpar}

To define a concrete extensional error ordering on interaction trees,
we define a type extensional execution traces. These traces are
finitary observations of an interaction tree, but ignoring the $\tau$
nodes.

We define two mutually inductive
types, $\textrm{PTrace}, \textrm{OTrace}$ which represent traces
starting with ``Player's'' turn or ``Opponent's'' turn, with
terminology inspired by game semantics. Here Player represents the
interaction tree, and the opponent represents the external system
interacting with the tree. Fix a type $V$ of final values.
\begin{mathpar}
  \textrm{PTrace} = V \uplus \{ \mho \} \uplus (\sigma : \Sigma) \times \textrm{OTrace}(\ar(\sigma))\\
  \textrm{OTrace}(R) = \{ \textrm{end} \} \uplus R \times \textrm{PTrace}
\end{mathpar}
A player trace is either terminated with a value, terminated with an
error or a pair of an operation $\sigma$ and an opponent trace, where
the opponent is expected to respond with the type of the arity of
$\sigma$. An opponent trace with response type $R$ is either empty,
marked by $\textrm{end}$ or it is a pair of a response $R$ and a
player trace. The role of the opponent trace is to allow for
observation of finite prefixes of infinitely long productive
sequences. Without it, the only observations would be of termination
behavior.

We write $t \downarrow \eta V$, $t \downarrow \mho$ and $t \downarrow
\textrm{end}$ to indicate that a player trace terminates in a
particular way.

We define by induction on the traces satisfaction
relations $M \vDash t$ between an ITree $M$ and a $\textrm{PTrace}$
$t$ and $f \vDash t$ between an $R$-indexed family of ITrees and an
$\textrm{OTrace}R$:
\begin{align*}
  {M \vDash \eta(V)} &= M \leadstomany \leaf(\eta(V))\\
  {M \vDash \mho} &= M \leadstomany \leaf(\mho)\\
  {M \vDash \sigma; o} &= M \leadstomany \sigma(f) \wedge f \vDash o\\
\end{align*}
\begin{align*}
  f \vDash \textrm{end} &= \top \\
  f \vDash r; p &= f(r) \vDash p
\end{align*}

Additionally, we inductively define an error ordering $\sqsubseteq$ on
player and opponent traces which intuitively means if $p \sqsubseteq q$ then $p$ is either
equal to $q$ or it is a prefix of $q$ that ends in $\mho$:
\begin{mathpar}
  \mho \sqsubseteq p

  \eta V \sqsubseteq \eta V

  \inferrule{o \sqsubseteq o'}{\sigma; o \sqsubseteq \sigma; o'}

  {\textrm{end} \sqsubseteq \textrm{end}}

  \inferrule{p \sqsubseteq q}{r; p \sqsubseteq r; q}
\end{mathpar}

Then the extensional error ordering can be defined as saying that $M
\ltbisim N$ when the results of observing $M$ are either the same as
$N$ or result in an error earlier on. 

\begin{align*}
   M \ltbisim N &= \forall (t : \textrm{PTrace}). \\
   &((N \vDash t) \Rightarrow \exists s \sqsubseteq t.~ M \vDash s)\\
   &\wedge ((M \vDash t) \wedge (t \downarrow \eta V \vee t \downarrow \textrm{end}) \Rightarrow N \vDash t)
\end{align*}

In plain English: if $N$ exhibits a finite trace $t$, then $M$ must
exhibit $t$ or an $\mho$-ending prefix of $t$, and if $M$ exhibits a
trace ending in a value or opponent's move, then $N$ must exhibit the
same trace.

\begin{lemma}[Properties of $\sqsubseteq$]
  \label{lem:itree-strong-err}
  Assume $M \sqsubseteq N$.
  \begin{enumerate}
  \item If $M \leadsto \eta x$ then $N \leadsto \eta x$
  \item If $M \leadsto \sigma(f)$ then $N \leadsto \sigma(g)$ with $\forall r\in \ar(\sigma). f(r) \sqsubseteq g(r)$.
  \item If $N \leadsto \err$ then $M \leadsto \err$
  \item If $N \leadsto \eta x$ then $M \leadsto \err$ or $M \leadsto \eta x$
  \item If $N \leadsto \sigma(f)$ then $M \leadsto \err$ or $M \leadsto \sigma(g)$ with $\forall r \in \ar(\sigma). f(r) \sqsubseteq g(r)$
  \end{enumerate}
\end{lemma}

\begin{lemma}[Properties of $\bisim$]
  \label{lem:itree-bisim}
  Assume $M \sqsubseteq N$.
  \begin{enumerate}
  \item If $M \leadsto \err$ then $N \leadsto \err$
  \item If $M \leadsto \eta x$ then $N \leadsto \eta x$
  \item If $M \leadsto \sigma(f)$ then $N \leadsto \sigma(g)$ with $\forall r\in \ar(\sigma). f(r) \bisim g(r)$.
  \item If $N \leadsto \err$ then $M \leadsto \err$
  \item If $N \leadsto \eta x$ then $M \leadsto \eta x$
  \item If $N \leadsto \sigma(f)$ then $M \leadsto \sigma(g)$ with $\forall r \in \ar(\sigma). f(r) \bisim g(r)$
  \end{enumerate}
\end{lemma}

\begin{theorem}[Properties of $\ltbisim$]
  $\ltbisim$ is transitive.
  If $M \sqsubseteq N$ then $M \ltbisim N$.
  If $M \bisim N$ then $M \ltbisim N$.
\end{theorem}
\begin{proof}
  Transitivity is clear. The other two implicications follow by induction on the structure of the input trace and the lemmas \ref{lem:itree-strong-err,lem:itree-bisim}
  By induction on the structure of the input trace.
\end{proof}

\bibliographystyle{ACM-Reference-Format}
\begin{DIFnomarkup}
\bibliography{\bibliographypath}
\end{DIFnomarkup}

% EXTENDED VERSION
\newpage
\appendix
% \input{appendix}

\end{document}
